### 2013-10-26 paper/main.tex
start, basic files in place

% {{{ preamble

%\documentclass[10pt,letterpaper]{article}
%\documentclass{article}
\documentclass[twocolumn]{article}
%\documentclass[12pt]{report}

\usepackage{url}
\usepackage{hyperref}
\usepackage{graphicx}
\usepackage{multicol}
\usepackage{nonfloat}
\usepackage{amsmath}
\usepackage{mathtools}

\usepackage{listings}
\lstset{numbers=left,
		language=Verilog,
%		basicstyle=\footnotesize,
		captionpos=b,
		xleftmargin=0.3in}

\usepackage{sectsty}  % \sectionfont
\sectionfont{\normalsize}
\subsectionfont{\normalsize}

\usepackage{vmargin}  % make the margins a bit smaller
\setmarginsrb{1.0in}{1.0in}{1.0in}{1.0in}{0in}{0.25in}{0in}{0.20in}

%\raggedright
\setlength{\parindent}{0.2in}

%
% backend	: biber
% style		: numeric
% autocite	: footnote
% citestyle	: verbose-inote
% bibstyle	: authortitle, numeric
%
\usepackage[backend=biber,autocite=footnote,
			bibstyle=authortitle,citestyle=verbose-inote]{biblatex}

%\addbibresource{main.bib}
\addbibresource{control.bib}

% }}}

\begin{document}

\twocolumn[\begin{@twocolumnfalse}

% {{{ title page

\thispagestyle{empty}

\centerline{\Large \textbf{A Survey of Control Systems Applied to}}
\centerline{\Large \textbf{the Idle Control of an Automotive Engine}}
\vspace{0.1in}
\centerline{\normalsize {Jeremiah Mahler}}
\centerline{\small {\href{mailto:jmahler@mail.csuchico.edu}{jmahler@mail.csuchico.edu}} }
\vspace{0.1in}
\centerline{\normalsize {CSU Chico}}
%\centerline{\today}
%\vspace{0.1in}
\centerline{\small \today}
\vspace{0.2in}
\centerline{\LARGE \textbf{DRAFT}}
\vspace{0.2in}

% }}}

\end{@twocolumnfalse}]

%\begin{multicols}{2}

% {{{ abstract
%\pagebreak
\thispagestyle{empty}
\begin{abstract}
\noindent

% TOOD

\end{abstract}
% }}}

\section{Model}

% TODO

\nocite{532315}

% References
\pagebreak
\printbibliography[heading=bibintoc]

\end{document}

### 2013-10-27 paper/main.tex
additional model reference
model with no controller working and documented

@@
 %\documentclass[10pt,letterpaper]{article}
-%\documentclass{article}
-\documentclass[twocolumn]{article}
+\documentclass{article}
+%\documentclass[twocolumn]{article}
 %\documentclass[12pt]{report}
@@
 \usepackage{graphicx}
-\usepackage{multicol}
+%\usepackage{multicol}
 \usepackage{nonfloat}
@@
 \usepackage{mathtools}
+\usepackage{parskip}
 
@@
 \lstset{numbers=left,
-		language=Verilog,
-%		basicstyle=\footnotesize,
+		language=Matlab,
+		basicstyle=\footnotesize,
 		captionpos=b,
@@
 
-%\raggedright
-\setlength{\parindent}{0.2in}
+\raggedright
+%\setlength{\parindent}{0.2in}
 
@@
 \begin{document}
-
-\twocolumn[\begin{@twocolumnfalse}
 
@@
 
-\end{@twocolumnfalse}]
-
-%\begin{multicols}{2}
-
 % {{{ abstract
 %\pagebreak
-\thispagestyle{empty}
-\begin{abstract}
-\noindent
+%\thispagestyle{empty}
+%\begin{abstract}
+%\noindent
 
@@
 
-\end{abstract}
+% What methods are used? PID, direct? ...
+
+%\end{abstract}
 % }}}
@@
 
-% TODO
+The engine model used here is based work
+by Butts and Sivashankar\autocite{532315} which is derived from
+the work by Powell and Cook\autocite{4789342}.
+The engine configuration is a modern 4.6L V-8.
+To simplify analysis the linearized model is used as shown in
+Figure \ref{fig:lem}.
 
-\nocite{532315}
+\begin{figure}[hbp!]
+\begin{center}
+\includegraphics[scale=0.6]{img/schematic-linear_engine_model-ed1}
+\end{center}
+\caption{Linear engine model.}\label{fig:lem}
+\end{figure}
+
+This model takes two inputs: a torque, and a idle control signal.
+When the torque is greater than zero it will oppose the rotation of
+the engine causing it to slow down.
+The idle control signal is some fraction of unity.
+This fraction corresponds to a pulse width modulated idle control valve
+which is at a minimum near zero and at a maximum near unity.
+Often the duty cycle range is in a range from 0\% to 100\% which
+corresponds to 0 to 1 (unity).
+
+This model has one complication that makes it difficult to use compared
+to typical elementary control systems.
+All the inputs and outputs are specified as deltas ($\Delta$).
+The torque input, for example, cannot be given as a constant ($T$) because this
+actually means $\Delta T$
+
+To overcome this complication a simple feedback controller can be used
+as show in Figure \ref{fig:scont}.
+And the open loop controller, with a fixed input duty cycle, is shown
+in Figure \ref{fig:openloop}.
+Notice that in this example the torque has transient characteristics and
+varies over time.
+The output of this system is show in Figure \ref{fig:olplot}.
+
+\clearpage
+\begin{figure}[!htbp]
+\begin{center}
+\includegraphics[scale=0.6]{img/schematic-control-ed1}
+\end{center}
+\caption{Simple feedback controller for inputs.
+The gain can be used to slow down the input changes.
+The output is an accumulation of all the step inputs.}
+\label{fig:scont}
+\end{figure}
+
+\begin{figure}[!htbp]
+\begin{center}
+\includegraphics[scale=0.5]{img/schematic-no_control-ed1}
+\end{center}
+\caption{Linear engine model with an open loop controller.
+Inputs go through the simple controller to provide steps to the
+engine model.
+The output rpm, since it is in steps ($\Delta N$), is accumulated.}
+\label{fig:openloop}
+\end{figure}
+
+\begin{figure}
+\begin{center}
+\includegraphics[scale=0.8]{img/linear_engine_model_no_control_plot}
+\end{center}
+\caption{Output of linear engine controller with a transient torque
+input and a fixed input duty cycle.  Matlab source given in
+Appendix \ref{app:olplot}}
+\label{fig:olplot}
+\end{figure}
 
 % References
-\pagebreak
+\clearpage
 \printbibliography[heading=bibintoc]
 
+\clearpage
+\appendix
+
+\section{Linear Engine Model Plot Matlab Script}
+\label{app:olplot}
+\lstinputlisting[language=Matlab]{../matlab/linear_engine_model_plot.m}
+
+\section{Linear Engine Model Initialization Matlab Script}
+\lstinputlisting[language=Matlab]{../matlab/linear_engine_model_init.m}
+
 \end{document}

### 2013-10-27 matlab/linear_engine_model_init.m

@@ -1,0 +1,16 @@
+% initialization for the Linear Engine Model
+% 'linear_engine_model.mdl'
+
+idle_rpm = 700;  % (RPM)
+num_cyl = 8;
+% frequency of ignition impulses
+% In two crankshaft rotations all cylinders will fire
+% for a 4-stroke engine.
+idle_rps = 700/60;  % minute -> second
+ign_freq = num_cyl*idle_rps*0.5;  % (Hz)
+
+% engine model time step
+Tm = 1/ign_freq;  % (sec)
+
+% control time step
+Tc = 0.5;  % (sec)

### 2013-10-27 matlab/linear_engine_model_plot.m

@@ -1,0 +1,34 @@
+clear all;
+clf;
+
+linear_engine_model_init;
+
+sim('linear_engine_model');
+
+figure;
+
+
+
+subplot(3,1,1);
+plot(dout.Time, dout.Data(:,1));
+grid on;
+axis tight;
+title('Idle Duty');
+ylabel('idle duty (%)');
+xlabel('time (sec)');
+
+subplot(3,1,2);
+plot(dout.Time, dout.Data(:,3));
+grid on;
+axis tight;
+title('Torque');
+ylabel('torque');
+xlabel('time (sec)');
+
+subplot(3,1,3);
+plot(dout.Time, dout.Data(:,2));
+grid on;
+axis tight;
+title('RPM');
+ylabel('RPM');
+xlabel('time (sec)');

### 2013-11-28 paper/main.tex
added some tikz diagrams, still a work in progress

@@
 \usepackage{parskip}
+\usepackage{bloques}
+
+\usepackage{tikz}
+\usetikzlibrary{calc,arrows,positioning}
+% control system tikz objects
+\tikzset{shadow/.style={drop shadow={opacity=0.8},fill=white}}
+\tikzset{sum/.style={circle,draw,very thick,shadow,minimum size=6mm}}
+\tikzset{sumnp/.style={sum,
+				label={{above left,xshift=1.0mm,yshift=-1.0mm}:$+$},
+				label={{below left,xshift=1.0mm,yshift=1.0mm}:$-$}}}
+\tikzset{sumxpxp/.style={sum,
+				label={{above left,xshift=1.0mm,yshift=-1.0mm}:$+$},
+				label={{above right,xshift=-1.0mm,yshift=-1.0mm}:$+$}}}
+\tikzset{gain/.style={rectangle,draw,very thick,inner sep=2mm,shadow}}
+%\newcommand{\nodesum}[2]{
+%\node[sum]  (#1) [label={{above left,xshift=1.5mm,yshift=-1.5mm}:$+$},
+%				  label={{below left,xshift=1.5mm,yshift=1.5mm}:$-$},#2] {};
+%}
 
@@
 
+\xdistance{1.8cm}
+\ydistance{2.0cm}
+
 \begin{figure}[hbp!]
 \begin{center}
-\includegraphics[scale=0.6]{img/schematic-linear_engine_model-ed1}
+\begin{tikzpicture}
+\bShadow
+\bStart{$\Delta \text{control}$}
+\bGain{$1.699$}
+\bMinusF{S1}
+\bGain{$\dfrac{8.5683}{z^2 - 0.9025 z}$}
+\bMinusUp{$\Delta \text{torque}$}
+\bGain{$\dfrac{2.98}{z - 0.9354}$}
+\bGainFeedBack{$0.000093$}{S1}
+\bEnd{$\Delta \text{rpm}$}
+\end{tikzpicture}
 \end{center}
@@
 This model has one complication that makes it difficult to use compared
-to typical elementary control systems.
+to typical control systems.
 All the inputs and outputs are specified as deltas ($\Delta$).
@@
 \begin{center}
-\includegraphics[scale=0.6]{img/schematic-control-ed1}
+
+\begin{tikzpicture}
+%\xdistance{1.2cm}
+%\ydistance{1.5cm}
+\bShadow
+\bStart{$X$}
+\bGain{$1 - z^{-1}$}
+\bEnd{$\Delta X$}
+\end{tikzpicture}
+
+%\begin{tikzpicture}
+%\xdistance{1.2cm}
+%\ydistance{1.5cm}
+%\bShadow
+%\bStart{$X$}
+%\bMinusF{S1}
+%\bGain{$\Sigma$}
+%\bFeedBack{}{S1}
+%\bEnd{$\Delta X$}
+%\end{tikzpicture}
+
 \end{center}
-\caption{Simple feedback controller for inputs.
-The gain can be used to slow down the input changes.
-The output is an accumulation of all the step inputs.}
+\caption{$Z$ transform used to accumulate the input and convert
+a steady state input to delta output.
+See Appendix \ref{app:cdelta} for the derivation.}
 \label{fig:scont}
@@
 
+% {{{ Steady State to Delta $Z$ Transform Derivation
+\section{Steady State to Delta $Z$ Transform Derivation}
+\label{app:cdelta}
+
+\begin{figure}
+\begin{center}
+
+\begin{tikzpicture}[>=triangle 60,
+	node distance=14mm, auto]
+%	[inner sep=10mm, minimum size=6mm, node distance=20mm]
+
+	\node[sumnp]  (sum1)	[] 					{};
+	\node[sumxpxp] (sum2)	[below=of sum1] 	{};
+	\node[gain] (gain1)		[right=of sum2]		{$z^{-1}$};
+	\node[gain] (gain2)		[left=of sum2]		{$z^{-1}$};
+
+	\draw [->] (sum2.north) -- 					(sum1.south);
+	\draw [->] (gain1.west) -- node[below]{$v$} 	(sum2.east);
+	\draw [->] (gain2.east) -- node[below]{$q$} 	(sum2.west);
+
+	 \draw [->]
+	 		($ (sum1.south) - (0,7mm) $)
+			-- ++(-35mm,0) node[below left]{$r$} 
+			|- (gain2.west);
+	\draw [->]
+			(-45mm,0) node[above] {$u$} to (sum1);
+	\draw [->]
+			(sum1) to (45mm,0) node[above] {$y$};
+	\draw [->]
+			($(sum1.east) + (33mm,0) $)
+			|- (gain1);
+
+\end{tikzpicture}
+
+%\tikzstyle{block}=[draw,minimum size=2.4em]
+%
+%\tikzstyle{sum}=[draw,circle,minimum size=1.2em]
+%
+%%\tikzstyle{init} = [pin edge={to-,thin,black}]
+%\begin{tikzpicture}[node distance=2.0cm,auto,>=latex']
+%
+%	% place all the block, points, relative to each other
+%	\node [coordinate,node distance=2cm] (u) {u};
+%	\node [block,right of=u] (k) {$K$};
+%
+%	\node [sum,right of=k] (s) {};
+%
+%	\node [coordinate,below of=s] (ds) {ds};
+%	\node [block,right of=ds] (d) {$D(s)$};
+%	\node [block,right of=s] (g) {$G(s)$};
+%	\node [coordinate,right of=g] (gdy) {};
+%	\node [coordinate,right of=gdy,node distance=2cm] (y) {y};
+%	\node [coordinate,right of=d] (gd) {gd};
+%
+%	% add signs to sum
+%	\node [left of=s,node distance=0cm,xshift=-3mm,yshift=3mm] {$+$};
+%	\node [left of=s,node distance=0cm,xshift=-3mm,yshift=-4mm] {$-$};
+%
+%	% add labels to G(s) and D(s)
+%	\node [right of=g,node distance=0cm,xshift=8mm,yshift=6mm] {$\dfrac{B}{A}$};
+%	\node [right of=d,node distance=0cm,xshift=8mm,yshift=6mm] {$\dfrac{\beta}{\alpha}$};
+%
+%	% connect all the nodes and points
+%	\path[->] (u) edge node {$u$} (k);
+%	\path[->] (k) edge node {} (s);
+%	\path[->] (ds) edge node {} (s);
+%	\path[-]  (d) edge node {} (ds);
+%	\path[->] (s) edge node {$\epsilon$} (g);
+%	\path[-] (g) edge node {} (gdy);
+%	\path[->] (gdy) edge node {$y$} (y);
+%	\path[-] (gdy) edge node {} (gd);
+%	\path[->] (gd) edge node {} (d);
+%
+%\end{tikzpicture}
+
+\end{center}
+
+\caption{System to accumulate values to convert a steady state
+input to a delta output.}
+\end{figure}
+
+% }}}
+
 \section{Linear Engine Model Plot Matlab Script}

### 2013-11-29 paper/main.tex
added steady state to delta derivation
All the engine model inputs are deltas so a transfer function
to has been found that takes a steady state input and produces
a delta output.  Then common control systems which produce
a steady state control can be used.

--- 2013-10-27 paper/main.tex
@@
 
-\usepackage{url}
+%\usepackage{url}
 \usepackage{hyperref}
 \usepackage{graphicx}
-%\usepackage{multicol}
 \usepackage{nonfloat}
@@
 \usepackage{parskip}
+\usepackage{bloques}
+\usepackage{fullpage}
+
+\usepackage{tikz}
+\usetikzlibrary{calc,arrows,positioning}
+% control system tikz objects
+\tikzset{shadow/.style={drop shadow={opacity=0.8},fill=white}}
+\tikzset{sum/.style={circle,draw,very thick,shadow,minimum size=6mm}}
+\tikzset{sumnp/.style={sum,
+				label={{above left,xshift=1.0mm,yshift=-1.0mm}:$+$},
+				label={{below left,xshift=1.0mm,yshift=1.0mm}:$-$}} }
+\tikzset{sumxpxp/.style={sum,
+				label={{above left,xshift=1.0mm,yshift=-1.0mm}:$+$},
+				label={{above right,xshift=-1.0mm,yshift=-1.0mm}:$+$}} }
+\tikzset{gain/.style={rectangle,draw,very thick,inner sep=2mm,shadow}}
 
@@
 		captionpos=b,
+		showspaces=false,
+		showstringspaces=false,
 		xleftmargin=0.3in}
@@
 \subsectionfont{\normalsize}
-
-\usepackage{vmargin}  % make the margins a bit smaller
-\setmarginsrb{1.0in}{1.0in}{1.0in}{1.0in}{0in}{0.25in}{0in}{0.20in}
 
@@
 
-%\addbibresource{main.bib}
-\addbibresource{control.bib}
+\addbibresource{main.bib}
+\setlength\bibitemsep{1em}
 
@@
 \begin{center}
-\includegraphics[scale=0.6]{img/schematic-control-ed1}
+
+\begin{tikzpicture}[>=triangle 60]
+\node[gain] (gain1) [] {$1 - z^{-1}$};
+\draw [->] (-30mm,0) node[above,yshift=1mm]{$X$} -- (gain1.west);
+\draw [->] (gain1.east) -- (30mm,0) node[above,yshift=1mm]{$\Delta X$};
+\end{tikzpicture}
+
 \end{center}
-\caption{Simple feedback controller for inputs.
-The gain can be used to slow down the input changes.
-The output is an accumulation of all the step inputs.}
+\caption{$Z$ transform used to accumulate the input and convert
+a steady state input to delta output.
+For its derivation, see Appendix \ref{app:cdelta}.}
 \label{fig:scont}
@@
 
+% {{{ Steady State to Delta $Z$ Transform Derivation
+\clearpage
+\section{Steady State to Delta $Z$ Transform Derivation}
+\label{app:cdelta}
+
+To accumulate a steady state input to produce a delta output
+a system can be constructed as shown in Figure \ref{fig:cd1}.
+Its operation can be confirmed by trying some values.
+If all values are zero and then a $1$ is input on $u$ the
+output will become $1$.
+On the next time step $1$ will be output on $v$.
+Since $q$ is zero $r$ will be $1$.
+If the input ($u$) remains $1$ this will be subtracted from $r$
+to produce zero on the output ($y$).
+
+% {{{ fig:cd1
+\begin{figure}[hpb!]
+\begin{center}
+
+\begin{tikzpicture}[>=triangle 60,
+	node distance=14mm, auto]
+
+	\node[sumnp]  	(sum1)	[] 					{};
+	\node[sumxpxp] 	(sum2)	[below=of sum1] 	{};
+	\node[gain] 	(gain1)	[right=of sum2]		{$z^{-1}$};
+	\node[gain] 	(gain2)	[left=of sum2]		{$z^{-1}$};
+
+	\draw [->] (sum2.north) -- 					(sum1.south);
+	\draw [->] (gain1.west) -- node[below]{$v$} 	(sum2.east);
+	\draw [->] (gain2.east) -- node[below]{$q$} 	(sum2.west);
+
+	 \draw [->]
+	 		($ (sum1.south) - (0,7mm) $)
+			-- ++(-35mm,0) node[below left,yshift=-3mm]{$r$} 
+			|- (gain2.west);
+	\draw [->]
+			(-45mm,0) node[above] {$u$} to (sum1);
+	\draw [->]
+			(sum1) to (45mm,0) node[above] {$y$};
+	\draw [->]
+			($(sum1.east) + (33mm,0) $)
+			|- (gain1);
+\end{tikzpicture}
+
+\end{center}
+
+\caption{System to accumulate values to convert a steady state
+input to a delta output.}
+\label{fig:cd1}
+\end{figure}
+
+% }}}
+
+Figure \ref{fig:cd1_plot} shows the response of this system given
+an arbitrary input.
+It can be seen that if the input is held constant the output (delta)
+returns to zero as expected.
+
+\begin{figure}[htbp!]
+\includegraphics[scale=0.7]{../octave/cd1_plot}
+\caption{Response of full steady state input to delta output system
+to an arbitrary input signal.
+The upper plot is the input signal ($u$) and the lower plot is
+the output response ($y$).
+The Matlab source code is given in Listing \ref{lst:cd1_init}
+and \ref{lst:cd1_plot}.
+}
+\label{fig:cd1_plot}
+\end{figure}
+
+\begin{samepage}
+However this full system can be simplified to a single transfer function.
+Starting from the equations that define the system
+\begin{align}
+	r &= q + v \label{eq:cd1a} \\
+	v &= y \cdot z^{-1} \label{eq:cd1b} \\
+	q &= r \cdot z^{-1} \label{eq:cd1c} \\
+	y &= u - r \label{eq:cd1d}
+\end{align}
+these can be algebraically manipulated to find the effective transfer
+function of the entire system ($y/u$).
+\end{samepage}
+
+\begin{align*}
+	r &= r z^{-1} + y z^{-1} && (\ref{eq:cd1a}, \ref{eq:cd1b}, \ref{eq:cd1c})\\
+	r(1 - z^{-1}) &= y z^{-1} \\
+	r &= u - y && (\ref{eq:cd1d} \\
+	(u - y)(1 - z^{-1}) &= y z^{-1} \\
+	u - y - u z^{-1} + y z^{-1} &= y z^{-1} \\
+	u - y - u z^{-1} &= 0 \\
+	y &= u (1 - z^{-1})
+\end{align*}
+\begin{align}
+	\Aboxed{ \frac{y}{u} &= 1 - z^{-1} }
+\end{align}
+
+\begin{figure}[!htbp]
+\begin{center}
+
+\begin{tikzpicture}[>=triangle 60]
+\node[gain] (gain1) [] {$1 - z^{-1}$};
+\draw [->] (-30mm,0) node[above,yshift=1mm]{$u$} -- (gain1.west);
+\draw [->] (gain1.east) -- (30mm,0) node[above,yshift=1mm]{$y$};
+\end{tikzpicture}
+
+\end{center}
+\caption{Simplified system to convert a steady state input in to
+a delta output.}
+\label{fig:cd1s}
+\end{figure}
+
+It can be seen in Figure \ref{fig:cd2_plot} that the simplified system
+behaves identically to the previous system (Figure \ref{fig:cd1_plot}).
+
+\begin{figure}[htbp!]
+\includegraphics[scale=0.7]{../octave/cd2_plot}
+\caption{Response of simplified steady state input to delta output system
+to an arbitrary input signal.
+The upper plot is the input signal ($u$) and the lower plot is
+the output response ($y$).
+Response is identical to the full system in Figure \ref{fig:cd1_plot}
+as expected.
+The Matlab source code is given in Listing \ref{lst:cd2_init}
+and \ref{lst:cd2_plot}.
+}
+\label{fig:cd2_plot}
+\end{figure}
+
+\clearpage
+\subsection{Matlab Source}
+\label{app:cdsrc}
+
+The following code has been tested using Octave\autocite{octave},
+an open source Matlab clone.
+
+\lstinputlisting[caption={Matlab code to initialize the full
+steady state to delta system.},
+	label=lst:cd1_init
+	]{../octave/cd1_init.m}
+
+\lstinputlisting[caption={Matlab code to plot the full steady state
+to delta system.},
+	label=lst:cd1_plot
+	]{../octave/cd1_plot.m}
+
+\clearpage
+\lstinputlisting[caption={Matlab code to initialize the simplified
+steady state to delta system.},
+	label=lst:cd2_init
+	]{../octave/cd2_init.m}
+
+\lstinputlisting[caption={Matlab code to plot the simplified steady state
+to delta system.},
+	label=lst:cd2_plot
+	]{../octave/cd2_plot.m}
+
+% }}}
+
+\clearpage
 \section{Linear Engine Model Plot Matlab Script}

### 2013-11-29 octave/cd2_init.m

@@ -1,0 +1,11 @@
+% 
+% cd2_init.m
+%
+% Convert a steady state input to a delta output.
+%
+% Simplified System.
+%
+
+T = 1;  % time step
+
+sys = tf([1 -1], [1 0], T, 'inname', 'y1', 'outname', 'u1');

### 2013-11-29 octave/cd1_init.m
added steady state to delta derivation
All the engine model inputs are deltas so a transfer function
to has been found that takes a steady state input and produces
a delta output.  Then common control systems which produce
a steady state control can be used.

@@ -1,0 +1,16 @@
+% 
+% cd1_init.m
+%
+% Convert a steady state input to a delta output.
+%
+% Full System.
+%
+
+T = 1;  % time step
+
+D1 = tf([1], [1 0], T, 'inname', 'y1', 'outname', 'v1');
+D2 = tf([1], [1 0], T, 'inname', 'r1', 'outname', 'q1');
+sum1 = sumblk('y1 = u1 - r1');
+sum2 = sumblk('r1 = v1 + q1');
+sys = connect(D1, D2, sum1, sum2, 'u1', 'y1');
+

### 2013-11-29 octave/cd1_plot.m

@@ -1,0 +1,25 @@
+%
+% cd1_plot.m
+%
+
+clear;
+
+cd1_init;
+
+u = [1 1 1 -3 -3 2 5 0];
+t = 0:(size(u,2)-1);  % start at zero
+
+figure;
+subplot(2,1,1);
+stairs(t,u);
+grid on;
+axis auto;
+title('Input Signal');
+ylabel('u');
+
+subplot(2,1,2);
+lsim(sys, u);
+grid on;
+axis auto;
+
+print('cd1_plot.eps', '-deps');

### 2013-11-29 octave/cd2_plot.m

@@ -1,0 +1,25 @@
+%
+% cd2_plot.m
+%
+
+clear;
+
+cd2_init;
+
+u = [1 1 1 -3 -3 2 5 0];
+t = 0:(size(u,2)-1);  % start at zero
+
+figure;
+subplot(2,1,1);
+stairs(t,u);
+grid on;
+axis auto;
+title('Input Signal');
+ylabel('u');
+
+subplot(2,1,2);
+lsim(sys, u);
+grid on;
+axis auto;
+
+print('cd2_plot.eps', '-deps');

### 2013-11-29 paper/main.tex
eliminated use of Bloques for TikZ directly
All uses of Bloques, a library for creating control system
diagrams, has been removed.  It did produce some nice graphics
but it was inflexible and difficult to fine tune.
Creating control system diagrams directly in TikZ along with
styles produces better results and is easier to maintain.

@@
 \usepackage{parskip}
-\usepackage{bloques}
 \usepackage{fullpage}
@@
 \usepackage{tikz}
-\usetikzlibrary{calc,arrows,positioning}
+\usetikzlibrary{calc,arrows,positioning,shadows}
 % control system tikz objects
-\tikzset{shadow/.style={drop shadow={opacity=0.8},fill=white}}
+\tikzset{shadow/.style={drop shadow,fill=white}}
 \tikzset{sum/.style={circle,draw,very thick,shadow,minimum size=6mm}}
@@
 				label={{below left,xshift=1.0mm,yshift=1.0mm}:$-$}} }
+\tikzset{sumpn/.style={sum,
+				label={{above left,xshift=1.0mm,yshift=-1.0mm}:$+$},
+				label={{above right,xshift=-1.0mm,yshift=-1.0mm}:$-$}} }
 \tikzset{sumxpxp/.style={sum,
@@
 
-\section{Model}
+\section{Engine Model}
 
@@
 
-\xdistance{1.8cm}
-\ydistance{2.0cm}
-
 \begin{figure}[hbp!]
 \begin{center}
-\begin{tikzpicture}
-\bShadow
-\bStart{$\Delta \text{control}$}
-\bGain{$1.699$}
-\bMinusF{S1}
-\bGain{$\dfrac{8.5683}{z^2 - 0.9025 z}$}
-\bMinusUp{$\Delta \text{torque}$}
-\bGain{$\dfrac{2.98}{z - 0.9354}$}
-\bGainFeedBack{$0.000093$}{S1}
-\bEnd{$\Delta \text{rpm}$}
+
+\begin{tikzpicture}[>=triangle 60, scale=0.8,
+	node distance=14mm, auto]
+
+	\node[gain] 	(gain1)	[]					{$1.699$};
+	\node[sumnp]  	(sum1)	[right=of gain1] 	{};
+	\node[gain] 	(gain2)	[right=of sum1]		{$\dfrac{8.5683}{z^2-0.9025z}$};
+	\node[sumpn]  	(sum2)	[right=of gain2] 	{};
+	\node[gain] 	(gain3)	[right=of sum2]		{$\dfrac{2.98}{z - 0.9534}$};
+	\node[gain] 	(gain4)	[below=of sum2]		{$0.000093$};
+
+	\draw [->] (-20mm,0) node[above]{$\Delta \text{control}$} -- (gain1.west);
+	\draw [->] (gain1) -- (sum1);
+	\draw [->] (sum1) -- (gain2);
+	\draw [->] (gain2) -- (sum2);
+	\draw [->] (sum2) -- (gain3);
+	\draw [->] (gain3.east) -- ++(20mm,0) node[above]{$\Delta \text{rpm}$};
+	\draw [->] ($(gain3.east) + (10mm,0) $) |- (gain4.east);
+	\draw [->] (gain4.west) -| (sum1.south);
+	\draw [<-] (sum2.north) -- ++(0,10mm) node[above]{$\Delta \text{torque}$};
 \end{tikzpicture}
+
 \end{center}

### 2013-11-29 paper/main.tex
documented delta to steady state conversion
The transfer function to convert delta values to
steady state has been derived and documented.

@@
 \usetikzlibrary{calc,arrows,positioning,shadows}
+%
 % control system tikz objects
@@
 \tikzset{sumnp/.style={sum,
-				label={{above left,xshift=1.0mm,yshift=-1.0mm}:$+$},
-				label={{below left,xshift=1.0mm,yshift=1.0mm}:$-$}} }
+				label={{above left,xshift=0.0mm,yshift=-1.0mm}:$+$},
+				label={{below right,xshift=-1.0mm,yshift=0.0mm}:$-$}} }
+\tikzset{sumpp/.style={sum,
+				label={{above left,xshift=0.0mm,yshift=-1.0mm}:$+$},
+				label={{below right,xshift=-1.0mm,yshift=0.0mm}:$+$}} }
 \tikzset{sumpn/.style={sum,
-				label={{above left,xshift=1.0mm,yshift=-1.0mm}:$+$},
-				label={{above right,xshift=-1.0mm,yshift=-1.0mm}:$-$}} }
+				label={{below left,xshift=0.0mm,yshift=1.0mm}:$+$},
+				label={{above right,xshift=-1.0mm,yshift=0.0mm}:$-$}} }
 \tikzset{sumxpxp/.style={sum,
-				label={{above left,xshift=1.0mm,yshift=-1.0mm}:$+$},
-				label={{above right,xshift=-1.0mm,yshift=-1.0mm}:$+$}} }
+				label={{above left,xshift=0.0mm,yshift=-1.0mm}:$+$},
+				label={{above right,xshift=-0.0mm,yshift=-1.0mm}:$+$}} }
 \tikzset{gain/.style={rectangle,draw,very thick,inner sep=2mm,shadow}}
@@
 \end{center}
-\caption{Linear engine model.}\label{fig:lem}
+\caption{Linear engine model of a modern 4.6L V-8.}\label{fig:lem}
 \end{figure}
@@
 
-This model has one complication that makes it difficult to use compared
-to typical control systems.
-All the inputs and outputs are specified as deltas ($\Delta$).
-The torque input, for example, cannot be given as a constant ($T$) because this
-actually means $\Delta T$
+Because all the inputs and outputs are defined as deltas ($\Delta$)
+this model cannot be used directly with typical control systems
+which use steady state values.
+It is possible convert these deltas to steady state equivalents.
+Figure \ref{fig:sdtf} shows the transfer function to convert steady
+state values to delta values.
+Figure \ref{fig:dstf} shows the transfer function to delta values
+to steady state values.
 
-To overcome this complication a simple feedback controller can be used
-as show in Figure \ref{fig:scont}.
-And the open loop controller, with a fixed input duty cycle, is shown
-in Figure \ref{fig:openloop}.
-Notice that in this example the torque has transient characteristics and
-varies over time.
-The output of this system is show in Figure \ref{fig:olplot}.
+% {{{ fig:sdtf
+\begin{figure}[!htbp]
 
-\clearpage
-\begin{figure}[!htbp]
 \begin{center}
-
 \begin{tikzpicture}[>=triangle 60]
 \node[gain] (gain1) [] {$1 - z^{-1}$};
-\draw [->] (-30mm,0) node[above,yshift=1mm]{$X$} -- (gain1.west);
-\draw [->] (gain1.east) -- (30mm,0) node[above,yshift=1mm]{$\Delta X$};
+\draw [->] (-20mm,0) node[above,yshift=1mm]{$X$} -- (gain1.west);
+\draw [->] (gain1.east) -- (20mm,0) node[above,yshift=1mm]{$\Delta X$};
 \end{tikzpicture}
+\end{center}
 
+\caption{The $Z$ transform used to accumulate the input and convert
+a steady state input to delta output.
+Its derivation is given in Appendix \ref{app:cdelta}.}
+\label{fig:sdtf}
+
+\end{figure}
+% }}}
+
+% {{{ fig:dstf
+\begin{figure}[!htbp]
+
+\begin{center}
+\begin{tikzpicture}[>=triangle 60]
+\node[gain] (gain1) [] {$\dfrac{1}{1 - z^{-1}}$};
+\draw [->] (-20mm,0) node[above,yshift=1mm]{$\Delta X$} -- (gain1.west);
+\draw [->] (gain1.east) -- (20mm,0) node[above,yshift=1mm]{$X$};
+\end{tikzpicture}
 \end{center}
-\caption{$Z$ transform used to accumulate the input and convert
-a steady state input to delta output.
-For its derivation, see Appendix \ref{app:cdelta}.}
-\label{fig:scont}
+
+\caption{The $Z$ transform used to convert delta input to a steady
+state output.
+Its derivation is given in Appendix \ref{app:dsderiv}.}
+\label{fig:dstf}
+
 \end{figure}
+% }}}
 
-\begin{figure}[!htbp]
-\begin{center}
-\includegraphics[scale=0.5]{img/schematic-no_control-ed1}
-\end{center}
-\caption{Linear engine model with an open loop controller.
-Inputs go through the simple controller to provide steps to the
-engine model.
-The output rpm, since it is in steps ($\Delta N$), is accumulated.}
-\label{fig:openloop}
-\end{figure}
+% TODO - delta to steady state
 
-\begin{figure}
-\begin{center}
-\includegraphics[scale=0.8]{img/linear_engine_model_no_control_plot}
-\end{center}
-\caption{Output of linear engine controller with a transient torque
-input and a fixed input duty cycle.  Matlab source given in
-Appendix \ref{app:olplot}}
-\label{fig:olplot}
-\end{figure}
+Typical control systems have an associated time step which can be
+varied and is crucial in determining the performance of
+the controller with regard to the Nyquist frequency.
+This model has no such constraint because one calculation step
+corresponds directly to one engine ignition event.
+The controller is effectively running at the same ``clock speed'' as
+the engine.
+
+% TODO - steady state model
+
+%And the open loop controller, with a fixed input duty cycle, is shown
+%in Figure \ref{fig:openloop}.
+%Notice that in this example the torque has transient characteristics and
+%varies over time.
+%The output of this system is show in Figure \ref{fig:olplot}.
+
+% TODO - open loop performance
+
+%\begin{figure}[!htbp]
+%\begin{center}
+%\includegraphics[scale=0.5]{img/schematic-no_control-ed1}
+%\end{center}
+%\caption{Linear engine model with an open loop controller.
+%Inputs go through the simple controller to provide steps to the
+%engine model.
+%The output rpm, since it is in steps ($\Delta N$), is accumulated.}
+%\label{fig:openloop}
+%\end{figure}
+
+%\begin{figure}
+%\begin{center}
+%\includegraphics[scale=0.8]{img/linear_engine_model_no_control_plot}
+%\end{center}
+%%\caption{Output of linear engine controller with a transient torque
+%input and a fixed input duty cycle.  Matlab source given in
+%Appendix \ref{app:olplot}}
+%\label{fig:olplot}
+%\end{figure}
 
@@
 
-% {{{ Steady State to Delta $Z$ Transform Derivation
+% {{{ Steady State to Delta Transform Derivation
 \clearpage
-\section{Steady State to Delta $Z$ Transform Derivation}
+\section{Steady State to Delta Transform Derivation}
 \label{app:cdelta}
@@
 \node[gain] (gain1) [] {$1 - z^{-1}$};
-\draw [->] (-30mm,0) node[above,yshift=1mm]{$u$} -- (gain1.west);
-\draw [->] (gain1.east) -- (30mm,0) node[above,yshift=1mm]{$y$};
+\draw [->] (-20mm,0) node[above,yshift=1mm]{$u$} -- (gain1.west);
+\draw [->] (gain1.east) -- (20mm,0) node[above,yshift=1mm]{$y$};
 \end{tikzpicture}
@@
 
+% {{{ Delta to Steady State Transform Derivation
+\clearpage
+\section{Delta to Steady State Transform Derivation}
+\label{app:dsderiv}
+
+To convert a delta input to a steady state output it should
+sum the history of values.
+Figure \ref{fig:dc1} shows the system.
+
+% {{{ fig:dc1
+\begin{figure}[hpb!]
+\begin{center}
+
+\begin{tikzpicture}[>=triangle 60,
+	node distance=14mm, auto]
+
+	\node[sumpp]  	(sum1)	[] 						{};
+	\node[gain] 	(gain1)	[below right=of sum1]	{$z^{-1}$};
+
+	\draw [->] (gain1.west) -| (sum1.south) node[left,yshift=-10mm]{$r$};
+	\draw [<-] (sum1.west) -- ++(-20mm,0) node[above]{$u$};
+	\draw [->] (sum1.east) -- (50mm,0) node[above]{$y$};
+	\draw [->] ($ (sum1.east) + (30mm,0) $)
+			|- (gain1.east);
+\end{tikzpicture}
+
+\end{center}
+
+\caption{System to convert delta input to steady state output.}
+\label{fig:dc1}
+\end{figure}
+% }}}
+
+This system can be simplified in to a single transfer function
+as given by Equation \ref{eq:dc2s} and shown in Figure \ref{fig:dc1s}.
+
+\begin{align}
+	y &= u + r \notag \\
+	r &= y \cdot z^{-1} \notag \\
+	y &= u + y z^{-1} \notag \\
+	y &= y(1 - z^{-1}) \notag \\
+	\Aboxed{ \frac{y}{u} &= \frac{1}{1 - z^{-1}} } \label{eq:dc2s}
+\end{align}
+
+% {{{ fig:dc1s
+\begin{figure}[!htbp]
+\begin{center}
+
+\begin{tikzpicture}[>=triangle 60]
+\node[gain] (gain1) [] {$1 - z^{-1}$};
+\draw [->] (-20mm,0) node[above,yshift=1mm]{$u$} -- (gain1.west);
+\draw [->] (gain1.east) -- (20mm,0) node[above,yshift=1mm]{$y$};
+\end{tikzpicture}
+
+\end{center}
+\caption{Simplified system to convert a delta input to
+a steady state output.}
+\label{fig:dc1s}
+\end{figure}
+% }}}
+
+Figure \ref{fig:dc2_plot} shows the response of this system given
+an arbitrary input
+\footnote{This input is actually the output of the steady state input
+to delta output given in Appendix \ref{app:cdelta}.
+They are equal and opposite as expected.}.
+It can be seen that the output is held in a steady state according
+to the delta inputs as expected.
+
+% {{{ fig:dc2_plot
+\begin{figure}[htbp!]
+\includegraphics[scale=0.7]{../octave/dc2_plot}
+\caption{Response of delta input to steady state output system
+when given an arbitrary input signal.
+The upper plot is the input signal ($u$) and the lower plot is
+the output response ($y$).
+The Matlab source code is given in Listing \ref{lst:dc2_init}
+and \ref{lst:dc2_plot}.
+}
+\label{fig:dc2_plot}
+\end{figure}
+% }}}
+
+\clearpage
+\subsection{Matlab Source}
+\label{app:dcsrc}
+
+The following code has been tested using Octave\autocite{octave},
+an open source Matlab clone.
+
+%\lstinputlisting[caption={Matlab code to initialize the full
+%steady state to delta system.},
+%	label=lst:dc1_init
+%	]{../octave/dc1_init.m}
+%
+%\lstinputlisting[caption={Matlab code to plot the full steady state
+%to delta system.},
+%	label=lst:dc1_plot
+%	]{../octave/dc1_plot.m}
+
+\lstinputlisting[caption={Matlab code to initialize the simplified
+steady state to delta system.},
+	label=lst:dc2_init
+	]{../octave/dc2_init.m}
+
+\lstinputlisting[caption={Matlab code to plot the simplified steady state
+to delta system.},
+	label=lst:dc2_plot
+	]{../octave/dc2_plot.m}
+
+% }}}
+
 \clearpage

### 2013-11-29 octave/dc2_plot.m

@@ -1,0 +1,27 @@
+%
+% dc2_plot.m
+%
+
+clear;
+
+dc2_init;
+
+u = [1 0 0 -4 0 5 3 3];
+t = 0:(size(u,2)-1);  % start at zero
+
+figure;
+subplot(2,1,1);
+stairs(t,u);
+grid on;
+%axis auto;
+axis([0 7 -6 6]);
+title('Input Signal');
+ylabel('u');
+
+subplot(2,1,2);
+lsim(sys, u);
+grid on;
+axis auto;
+axis([0 7 -4 6]);
+
+print('dc2_plot.eps', '-deps');

### 2013-11-29 octave/dc1_init.m
documented delta to steady state conversion
The transfer function to convert delta values to
steady state has been derived and documented.

@@ -1,0 +1,13 @@
+% 
+% dc1_init.m
+%
+% Convert a delta input to a steady state output.
+%
+% Full System.
+%
+
+T = 1;  % time step
+
+D1 = tf([1], [1 0], T, 'inname', 'y1', 'outname', 'r1');
+S1 = sumblk('y1 = u1 + r1');
+sys = connect(D1, S1, 'u1', 'y1');

### 2013-11-29 octave/dc1_plot.m

@@ -1,0 +1,25 @@
+%
+% dc1_plot.m
+%
+
+clear;
+
+dc1_init;
+
+u = [1 0 0 -4 0 5 3 3];
+t = 0:(size(u,2)-1);  % start at zero
+
+figure;
+subplot(2,1,1);
+stairs(t,u);
+grid on;
+axis auto;
+title('Input Signal');
+ylabel('u');
+
+subplot(2,1,2);
+lsim(sys, u);
+grid on;
+axis auto;
+
+print('dc1_plot.eps', '-deps');

### 2013-11-29 octave/dc2_init.m

@@ -1,0 +1,11 @@
+% 
+% dc1_init.m
+%
+% Convert a delta input to a steady state output.
+%
+% Simplified System.
+%
+
+T = 1;  % time step
+
+sys = tf([1 0], [1 -1], T, 'inname', 'u1', 'outname', 'y1');

### 2013-11-30 paper/main.tex
more work on the engine model
Still trying to sort out the engine model so it can
be used with a controller.

@@
 
-% TODO - delta to steady state
+Typical control systems have an associated time step.
+And the choice of this time step is crucial in determining performance
+with regard to the Nyquist frequency.
+However this model does not suffer from this issue because it is
+inherently discrete.
+A single ignition event of the engine corresponds to a single
+step of the model.
 
-Typical control systems have an associated time step which can be
-varied and is crucial in determining the performance of
-the controller with regard to the Nyquist frequency.
-This model has no such constraint because one calculation step
-corresponds directly to one engine ignition event.
-The controller is effectively running at the same ``clock speed'' as
-the engine.
+% TODO - reduced steady state system
+
+Typical control system models have a single input and a single output
+\footnote{One input/output could also be a vector.}.
+However this model is different in that it takes
+two inputs: control and torque (Figure \ref{fig:lem}).
+One way to reduce the model to a single input single output is to hold
+one of the inputs constant.
+However neither of the resulting use cases would make any sense to
+a designer whose primary goal is to stabilize engine rpm.
+The first case, with torque constant, would indicate the response for
+control inputs.
+But this is pointless because the goal of idle stabilization is to
+regulate the rpm given changes in torque.
+The second case, with control held constant, would indicate the response
+for a change in torque.
+But again this would provide no useful way to stabilize engine rpm.
+Both inputs are needed in order to construct any viable control system.
+
+\section{Regulator Control System}
+
+What is needed is a regulator system where the input is applied
+as torque.
+The general system is shown in Figure \ref{fig:rpm1}.
+
+% {{{ engine regulator system
+\begin{figure}[hbp!]
+\begin{center}
+
+\begin{tikzpicture}[>=triangle 60, scale=0.8,
+	node distance=14mm, auto]
+
+	\node[gain] 	(engine)	[]					{engine};
+	\node[gain] 	(control)	[below=of engine]	{controller};
+
+	\draw [<-] (engine.north) -- ++(0,10mm) node[above]{$\text{torque}$};
+	\draw [->] (engine.east) node[above,xshift=7mm]{rpm}
+					-- ++(20mm,0)
+					|- (control.east);
+	\draw [->] (control.west)
+					-- ++(-20mm,0)
+					|- (engine.west) node[above,xshift=-9mm]{control};
+\end{tikzpicture}
+
+\end{center}
+\caption{General system for regulating engine rpm.}\label{fig:rpm1}
+\end{figure}
+% }}}
 
@@
 % TODO - open loop performance
-
-%\begin{figure}[!htbp]
-%\begin{center}
-%\includegraphics[scale=0.5]{img/schematic-no_control-ed1}
-%\end{center}
-%\caption{Linear engine model with an open loop controller.
-%Inputs go through the simple controller to provide steps to the
-%engine model.
-%The output rpm, since it is in steps ($\Delta N$), is accumulated.}
-%\label{fig:openloop}
-%\end{figure}
-
-%\begin{figure}
-%\begin{center}
-%\includegraphics[scale=0.8]{img/linear_engine_model_no_control_plot}
-%\end{center}
-%%\caption{Output of linear engine controller with a transient torque
-%input and a fixed input duty cycle.  Matlab source given in
-%Appendix \ref{app:olplot}}
-%\label{fig:olplot}
-%\end{figure}
 

### 2013-11-30 octave/delta_engine_model.m

@@ -1,0 +1,31 @@
+% 
+% delta_engine_model.m
+%
+% Linear engine model with delta inputs/outputs.
+%
+
+T = 1;  % time step
+
+D1 = tf([1.699], [1], T, 'inname', 'delta_control', 'outname', 'y1');
+S1 = sumblk('e1 = y1 - y3');
+D2 = tf([8.5683], [1 -0.9025 0], T, 'inname', 'e1', 'outname', 'y2');
+S2 = sumblk('e2 = y2 - delta_torque');
+D3 = tf([2.98], [1 -0.9534], T, 'inname', 'e2', 'outname', 'delta_rpm');
+D4 = tf([0.000093], [1], T, 'inname', 'delta_rpm', 'outname', 'y3');
+
+delta_engine_model_sys = connect(D1, D2, D3, D4, S1, S2, {'delta_control', 'delta_torque'}, 'delta_rpm');
+
+t_end = 100;
+Nt = t_end + 1 ;
+t = linspace(0, t_end, Nt);
+m = 2;
+%u = ones(Nt, m);  % step on all inputs
+u = [0*ones(Nt,1) 0*ones(Nt,1)];
+u(1,2) = 1;
+u(2,2) = 1;
+u(3,2) = 1;
+u(4,2) = 1;
+
+[y,t,x] = lsim(delta_engine_model_sys, u, t);
+
+plot(t,y);

### 2013-11-30 paper/main.tex
engine model with no control working
With the torque input is set to zero a transfer function
representing the engine model can be constructed.
This should be suitable for building control systems
using pole placements or other methods.

@@
 
-\begin{tikzpicture}[>=triangle 60, scale=0.8,
+\begin{tikzpicture}[>=triangle 60, scale=0.7,
 	node distance=14mm, auto]
@@
 	\node[sumpn]  	(sum2)	[right=of gain2] 	{};
-	\node[gain] 	(gain3)	[right=of sum2]		{$\dfrac{2.98}{z - 0.9534}$};
-	\node[gain] 	(gain4)	[below=of sum2]		{$0.000093$};
+	\node[gain] 	(gain3)	[right=of sum2]		{$\dfrac{2.98}{z - 0.9354}$};
+	\node[gain] 	(gain4)	[below=of sum2]		{$0.00093$};
 
@@
 
-Typical control system models have a single input and a single output
-\footnote{One input/output could also be a vector.}.
-However this model is different in that it takes
-two inputs: control and torque (Figure \ref{fig:lem}).
-One way to reduce the model to a single input single output is to hold
-one of the inputs constant.
-However neither of the resulting use cases would make any sense to
-a designer whose primary goal is to stabilize engine rpm.
-The first case, with torque constant, would indicate the response for
-control inputs.
-But this is pointless because the goal of idle stabilization is to
-regulate the rpm given changes in torque.
-The second case, with control held constant, would indicate the response
-for a change in torque.
-But again this would provide no useful way to stabilize engine rpm.
-Both inputs are needed in order to construct any viable control system.
+In order to build a controller the model needs to be simplified
+but the presence of two inputs complicates matters.
+To resolve this issue the torque can be set zero.
+With this change it can be simplified by first recognizing that
+it matches the form shown in Figure \ref{fig:dd} which
+has the transfer function in Equation \ref{eq:dd}.
 
+% {{{ dd figure
+\begin{figure}[h!]
+
+\begin{center}
+\begin{tikzpicture}[>=triangle 60, scale=0.8,
+	node distance=20mm, auto]
+
+	\node[gain]		(K) 	[] 				{$K$};
+	\node[sumnp]	(sum1) 	[right of=K] 	{};
+	\node[gain]		(Gz) 	[right of=sum1]	{$G(z)$};
+	\node[gain]		(Dz) 	[below of=Gz] 	{$D(z)$};
+
+	\draw [<-] (K) -- ++(-20mm,0) node[above]{$u$};
+	\draw [->] (K) -- (sum1);
+	\draw [->] (sum1) -- (Gz);
+	\draw [<-] (sum1) |- (Dz);
+	\draw [->] (Gz) -- ++(35mm,0) node[above]{$y$};
+	\draw [->] (Gz) -- ++(20mm,0) |- (Dz.east);
+\end{tikzpicture}
+\end{center}
+
+\caption{Direct Design system with $K$ as a scaling input, $G$ is
+the plant and $D$ is the controller.}
+\label{fig:dd}
+\end{figure}
+% }}}
+
+\begin{align}
+\frac{y}{u} &= K \frac{B \alpha}{A \alpha + B \beta} \label{eq:dd}
+\end{align}
+
+First taking the parts from the engine model (Figure \ref{fig:lem}).
+
+\begin{align*}
+	\frac{B}{A} &= \frac{8.5683}{(z^2 - 0.9025z)} \frac{2.98}{(z - 0.9354)} \\
+	K &= 1.699 \\
+	\frac{\beta}{\alpha} &= 0.00093 
+\end{align*}
+
+Then substituting them in to Equation \ref{eq:dd} results in
+the simplified from of Equation \ref{eq:lems}.
+
+\begin{align}
+	\frac{y}{u} &= \frac{(1.699) (8.5683) (2.98) (1)}
+		{(z^2 - 0.9025z)(z - 0.9354)(1) + (8.5683)(2.98)(0.00093)} \notag \\
+	\Aboxed{ \frac{y}{u} &= \frac{43.38}{z^3 - 1.838 z^2 + 0.8442 z + 0.02375}}
+		\label{eq:lems}
+\end{align}
+
+\begin{samepage}
+It is still necessary to address the delta inputs and outputs.
+It can be seen in Figure \ref{fig:nodelta} there is a beneficial
+side effect that makes this trivial.  The transform that converts
+from steady state to a delta cancels
+with the transform that converts from a delta to steady state
+\footnote{If the torque input was non zero the transforms would not
+cancel.}.  Therefore the final transform is still Equation \ref{eq:lems}.
+An example of the output response with no control is shown in
+Figure \ref{fig:em_plot}.
+\end{samepage}
+
+% {{{ fig:nodelta
+\begin{figure}[!htbp]
+
+\begin{center}
+\begin{tikzpicture}[>=triangle 60,
+	node distance=30mm, auto]
+
+\node[gain] (gain1) [] {$1 - z^{-1}$};
+\node[gain] (engine) [right of=gain1] {engine};
+\node[gain] (gain3) [right of=engine] {$\dfrac{1}{1 - z^{-1}}$};
+
+\draw [->] (-25mm,0) node[above]{$u$} -- (gain1.west);
+\draw [->] (gain1.east) node[above,xshift=6mm]{$\Delta u$} -- (engine);
+\draw [->] (engine.east) node[above,xshift=6mm]{$\Delta y$} -- (gain3);
+\draw [->] (gain3.east) -- ++(20mm,0) node[above]{$y$};
+\end{tikzpicture}
+\end{center}
+
+\caption{Engine model with transforms for converting from steady state
+to delta and vice versa.}
+\label{fig:nodelta}
+
+\end{figure}
+% }}}
+
+\begin{figure}[htbp!]
+\begin{center}
+\includegraphics[scale=0.7]{../octave/engine_model_plot}
+\end{center}
+\caption{Output response of engine model with no control.
+The control can range from zero to one.}
+Here the control is constant at 0.65 which resulted in a stable
+rpm near 900.
+The Matlab script is provided in Appendix \ref{app:em_matlab}.
+\label{fig:em_plot}
+\end{figure}
+
+% see octave/engine_model.m for these calculations
+
+\clearpage
 \section{Regulator Control System}
@@
 % TODO - steady state model
-
-%And the open loop controller, with a fixed input duty cycle, is shown
-%in Figure \ref{fig:openloop}.
-%Notice that in this example the torque has transient characteristics and
-%varies over time.
-%The output of this system is show in Figure \ref{fig:olplot}.
 
@@
 	y &= u + y z^{-1} \notag \\
-	y &= y(1 - z^{-1}) \notag \\
+	u &= y(1 - z^{-1}) \notag \\
 	\Aboxed{ \frac{y}{u} &= \frac{1}{1 - z^{-1}} } \label{eq:dc2s}
@@
 \begin{tikzpicture}[>=triangle 60]
-\node[gain] (gain1) [] {$1 - z^{-1}$};
+	\node[gain] (gain1) [] {$\dfrac{1}{1 - z^{-1}}$};
 \draw [->] (-20mm,0) node[above,yshift=1mm]{$u$} -- (gain1.west);
@@
 \clearpage
-\section{Linear Engine Model Plot Matlab Script}
-\label{app:olplot}
-\lstinputlisting[language=Matlab]{../matlab/linear_engine_model_plot.m}
+\section{Engine Model Matlab Source}
+\label{app:em_matlab}
 
-\section{Linear Engine Model Initialization Matlab Script}
-\lstinputlisting[language=Matlab]{../matlab/linear_engine_model_init.m}
+\lstinputlisting[
+	caption={Matlab source to calculate the engine model system with
+		zero torque.}
+	]{../octave/engine_model.m}
+
+\lstinputlisting[
+	caption={Matlab source to plot the response of engine model
+		with no control.}
+	]{../octave/engine_model_plot.m}
 

### 2013-11-30 octave/engine_model.m

@@ -1,0 +1,17 @@
+% 
+% engine_model.m
+%
+% Linear engine model with zero torque.
+%
+
+T = 1;  % time step
+
+K = tf([1.699], [1], T);
+D1= tf([8.5683], [1 -0.9025 0], T);
+D2= tf([2.98], [1 -0.9354], T);
+G = D1*D2;
+D = tf([0.00093], [1], T);
+
+H = K*G/(1 + D*G);
+
+eng_sys = minreal(H);  % cancel common roots

### 2013-11-30 octave/engine_model_plot.m

@@ -1,0 +1,28 @@
+%
+% engine_model_plot.m
+%
+
+clear;
+
+engine_model;
+
+u = [0.65*ones(1,100)];
+[y, t, x] = lsim(eng_sys, u);
+
+figure;
+subplot(2,1,1);
+
+stairs(t, u);
+grid on;
+axis auto;
+title('Input Signal');
+ylabel('control');
+
+subplot(2,1,2);
+stairs(t, y);
+grid on;
+axis auto;
+title('Output Response');
+ylabel('rpm');
+
+print('engine_model_plot.eps', '-deps');

### 2013-11-30 octave/delta_engine_model.m
engine model with no control working
With the torque input is set to zero a transfer function
representing the engine model can be constructed.
This should be suitable for building control systems
using pole placements or other methods.

@@ -1,31 +1,0 @@
-% 
-% delta_engine_model.m
-%
-% Linear engine model with delta inputs/outputs.
-%
-
-T = 1;  % time step
-
-D1 = tf([1.699], [1], T, 'inname', 'delta_control', 'outname', 'y1');
-S1 = sumblk('e1 = y1 - y3');
-D2 = tf([8.5683], [1 -0.9025 0], T, 'inname', 'e1', 'outname', 'y2');
-S2 = sumblk('e2 = y2 - delta_torque');
-D3 = tf([2.98], [1 -0.9534], T, 'inname', 'e2', 'outname', 'delta_rpm');
-D4 = tf([0.000093], [1], T, 'inname', 'delta_rpm', 'outname', 'y3');
-
-delta_engine_model_sys = connect(D1, D2, D3, D4, S1, S2, {'delta_control', 'delta_torque'}, 'delta_rpm');
-
-t_end = 100;
-Nt = t_end + 1 ;
-t = linspace(0, t_end, Nt);
-m = 2;
-%u = ones(Nt, m);  % step on all inputs
-u = [0*ones(Nt,1) 0*ones(Nt,1)];
-u(1,2) = 1;
-u(2,2) = 1;
-u(3,2) = 1;
-u(4,2) = 1;
-
-[y,t,x] = lsim(delta_engine_model_sys, u, t);
-
-plot(t,y);

### 2013-11-30 paper/main.tex
added abstract, minor cleanups

@@
 %\thispagestyle{empty}
-%\begin{abstract}
-%\noindent
+\begin{abstract}
+\noindent
+The task of maintaining a stable idle for an internal combustion
+engine with spark ignition is non-trivial.
+Any time an accessory is turned on/off the torque applied to
+the engine changes.
+And changes in torque will change the engine rpm if the control
+inputs are constant.
+This paper shows how control systems methods can be applied to
+the problem of idle stabilization.
+Methods include: pole placement, PID, direct design, and
+various state space designs.
 
-% TOOD
-
-% What methods are used? PID, direct? ...
-
-%\end{abstract}
+\end{abstract}
 % }}}
@@
 The engine model used here is based work
-by Butts and Sivashankar\autocite{532315} which is derived from
+by Butts and Sivashankar\autocite{532315} which was derived from
 the work by Powell and Cook\autocite{4789342}.
-The engine configuration is a modern 4.6L V-8.
+The configuration is a modern 4.6L V-8 gas engine.
+A single control input representing the idle air bypass valve
+
 To simplify analysis the linearized model is used as shown in
@@
 
+\begin{samepage}
 Then substituting them in to Equation \ref{eq:dd} results in
 the simplified from of Equation \ref{eq:lems}.
-
 \begin{align}
@@
 \end{align}
+\end{samepage}
 
@@
 It is still necessary to address the delta inputs and outputs.
-It can be seen in Figure \ref{fig:nodelta} there is a beneficial
-side effect that makes this trivial.  The transform that converts
-from steady state to a delta cancels
+It can be seen in Figure \ref{fig:nodelta} that there is a beneficial
+side effect which makes this trivial.
+The transform that converts from steady state to a delta cancels
 with the transform that converts from a delta to steady state
-\footnote{If the torque input was non zero the transforms would not
-cancel.}.  Therefore the final transform is still Equation \ref{eq:lems}.
+\footnote{In the more general case, when torque is non zero,
+the transforms would not cancel.}.
+Therefore the final transform is still Equation \ref{eq:lems}.
 An example of the output response with no control is shown in
@@
 \caption{Output response of engine model with no control.
-The control can range from zero to one.}
+The control can range from zero to one.
 Here the control is constant at 0.65 which resulted in a stable
-rpm near 900.
-The Matlab script is provided in Appendix \ref{app:em_matlab}.
+rpm near 900.\protect\footnotemark}
 \label{fig:em_plot}
 \end{figure}
-
-% see octave/engine_model.m for these calculations
+\footnotetext{The Matlab source to produced these plots is in Appendix \ref{app:em_matlab}.}
 

### 2013-12-01 octave/engine_model_rpmtime_plot.m
converted ignition points to time in model description
Previously the engine model response with no control
showed rpm versus ignition points.  But this did not
indicate how much time it took to stabilize.
A conversion was found to plot these with time.

@@ -1,0 +1,44 @@
+%
+% engine_model_rpmtime_plot.m
+%
+
+clear;
+
+ncyl = 8;
+
+engine_model;
+
+u = [0.65*ones(1,100)];
+[rpm, t, x] = lsim(eng_sys, u);
+
+% An rpm of zero makes the rpmtime huge and
+% skews the results.
+% Remove these initial zeros.
+i = find(rpm~=0, 1, 'first');
+rpm = rpm(i:end);
+u = u(i:end);
+t = t(i:end);
+
+% Convert each rpm point to time instant.
+% The cummulative sum is then the time it takes
+% to get to that particular point.
+rt = rpmtime(rpm, 8);
+rt = cumsum(rt);
+
+figure;
+subplot(2,1,1);
+stairs(rt, u);
+grid on;
+axis auto;
+title('Input Signal');
+ylabel('control');
+
+subplot(2,1,2);
+stairs(rt, rpm);
+grid on;
+axis auto;
+title('Output Response');
+ylabel('rpm');
+xlabel('time (sec)');
+
+print('engine_model_rpmtime_plot.eps', '-deps');

### 2013-12-01 paper/main.tex
converted ignition points to time in model description
Previously the engine model response with no control
showed rpm versus ignition points.  But this did not
indicate how much time it took to stabilize.
A conversion was found to plot these with time.

@@
 
-\centerline{\Large \textbf{A Survey of Control Systems Applied to}}
-\centerline{\Large \textbf{the Idle Control of an Automotive Engine}}
+\centerline{\Large \textbf{A Survey of Control Systems for Stabilizing}}
+\centerline{\Large \textbf{the Idle of a Spark Ignition Engine}}
 \vspace{0.1in}
@@
 
+% {{{ Engine Model
 \section{Engine Model}
@@
 
+% {{{ fig:em_plot
 \begin{figure}[htbp!]
@@
 \footnotetext{The Matlab source to produced these plots is in Appendix \ref{app:em_matlab}.}
+% }}}
 
+% }}}
+
+It is not apparent from Figure \ref{fig:em_plot} exactly how long
+this process takes.
+The $x$ axis represents ignition events, not time.
+The rpm can be converted to time per ignition event by examining
+the units as shown in Equation \ref{eq:rpmtime}.
+
+\begin{align}
+\frac{1}{x \frac{\text{rev}}{\text{min}}
+			\cdot n \frac{\text{ign}}{\text{rev}}
+			\cdot \frac{\text{min}}{60\text{sec}}
+		}
+&= y \frac{\text{sec}}{\text{ign}} \label{eq:rpmtime} \\
+& \notag \\
+	\frac{\text{rev}}{\text{min}} &: \text{revolutions per minute (rpm)} \notag \\
+	\frac{\text{ign}}{\text{rev}} &: \text{number ignition points per revolution (4 for 8 cylinder engine)} \notag \\
+	\frac{\text{sec}}{\text{ign}} &: \text{seconds per ignition point} \notag
+\end{align}
+
+Applying this equation results in the response shown
+in Figure \ref{fig:rpmtime_plot}.
+It can be seen that this plot reveals that it taks over two seconds
+for the system to stabilize with no control.
+Clearly this is unacceptably slow and it is likely that with
+the application of a controller its performance can be improved.
+
+% {{{ fig:rpmtime_plot
+\begin{figure}[htbp!]
+\begin{center}
+\includegraphics[scale=0.7]{../octave/engine_model_rpmtime_plot}
+\end{center}
+\caption{Engine rpm response compared to time with a constant
+0.65 input control.\protect\footnotemark}
+\label{fig:rpmtime_plot}
+\end{figure}
+\footnotetext{The Matlab source of the RPM vs time plot is in Appendix \ref{app:rpmtime_plot_source}.}
+% }}}
+
+\clearpage
+
+% {{{ Pole Placement
+\section{Pole Placement}
+
+% }}}
+
+% {{{ Regulator Control System
 \clearpage
@@
 
-% {{{ engine regulator system
+% {{{ fig:rpm1, Regulator System
 \begin{figure}[hbp!]
@@
 
-% TODO - steady state model
-
-% TODO - open loop performance
+% }}}
 
@@
 \label{app:em_matlab}
+\label{app:rpmtime_plot_source}
 
@@
 
+\clearpage
+\lstinputlisting[
+	caption={Matlab source to plot the response of engine model
+		versus time.}
+	]{../octave/engine_model_rpmtime_plot.m}
+
 \end{document}

### 2013-12-01 octave/rpmtime.m

@@ -1,0 +1,25 @@
+function [tign] = rpmtime(rpm, ncyl)
+% RPMTIME Calculate the time (in seconds) between ignition
+%  events at a given rpm.
+%
+%  rpmtime(rpm, ncyl);
+%  rpmtime(400, 8);
+%
+
+	%ncyl;				% number of cylinders
+	%rpm;				% rpm
+	ign_rev = ncyl/2;	% igntion points per revolution
+	min_sec = 1/60;		% 1 minute per 60 seconds
+	%tign;				% seconds per ignition event
+
+	tign = zeros(0,length(rpm));
+	for i = 1:length(rpm)
+		x = (rpm(i)*min_sec*ign_rev);
+		if (0 == x)
+			tign(i) = inf;
+		else
+			tign(i) = 1/x;
+		endif
+	endfor
+
+endfunction

### 2013-12-01 paper/main.tex
normalization corrections for engine model
Corrections to normalize the engine model have been found
which will accept a step input and produce a response near one.

@@
 To resolve this issue the torque can be set zero.
-With this change it can be simplified by first recognizing that
-it matches the form shown in Figure \ref{fig:dd} which
+Then it can be simplified by recognizing that it matches
+the well known form shown in Figure \ref{fig:dd} which
 has the transfer function in Equation \ref{eq:dd}.
@@
 \begin{samepage}
-Then substituting them in to Equation \ref{eq:dd} results in
-the simplified from of Equation \ref{eq:lems}.
+Then substituting them in to Equation \ref{eq:dd} and simplifying
+results in Equation \ref{eq:lems}.
 \begin{align}
@@
 It is still necessary to address the delta inputs and outputs.
-It can be seen in Figure \ref{fig:nodelta} that there is a beneficial
-side effect which makes this trivial.
+This can be resolved by placing the conversion transforms,
+from Figure \ref{fig:sdtf} and \ref{fig:dstf}, on either
+end of the engine model transfer function as shown in Figure \ref{fig:nodelta}.
+A beneficial side effect becomes apparent in this form.
 The transform that converts from steady state to a delta cancels
@@
 
-% }}}
-
 It is not apparent from Figure \ref{fig:em_plot} exactly how long
@@
 in Figure \ref{fig:rpmtime_plot}.
-It can be seen that this plot reveals that it taks over two seconds
-for the system to stabilize with no control.
-Clearly this is unacceptably slow and it is likely that with
-the application of a controller its performance can be improved.
+It can be seen it takes over two seconds for the system to
+stabilize with no control.
+Clearly this is unacceptably slow and performance could be
+improved by using a controller.
 
@@
 \footnotetext{The Matlab source of the RPM vs time plot is in Appendix \ref{app:rpmtime_plot_source}.}
+% }}}
+
+There is yet another complication with this model.
+Typical systems are in a normalized so that it is possible to
+achieve an output of one given a step input.
+With this engine model an output of 1 rpm is an impossibly slow value.
+And an input of zero or one rpm are similarly impossible.
+
+To resolve this issue the values will be scaled in to a normal range.
+A normal idle rpm range is from 600 to 1200 rpm.
+To scale these values so that zero equals 600 and one equals 1000
+Equation \ref{eq:scale} can be used.
+
+\begin{align}
+	y &= \frac{1}{400}x - \frac{3}{2} \label{eq:scale}
+\end{align}
+
+Figure \ref{fig:stable} shows the stable rpms for a given control input.
+Its behavior is described by Equation \ref{eq:stable}.
+
+\begin{align}
+	y &= 1450.1 x \label{eq:stable}
+\end{align}
+
+% {{{ fig:stable
+\begin{figure}[htbp!]
+\begin{center}
+\includegraphics[scale=0.6]{../octave/stable_rpm_plot}
+\end{center}
+\caption{Stable rpm reached for a given constant control input.}
+\label{fig:stable}
+\end{figure}
+% }}}
+
+With these normalization corrections applied the response can be
+seen in Figure \ref{fig:normeng}.
+To avoid the zero rpm case it is allowed to stabilize before the
+step is applied.
+The lowest plot shows the response after the step.
+It takes slightly over one second to stabilize, which is better than
+the inaccurate case from zero rpm, but it is still slow and can
+be improved.
+
+% {{{ fig:normeng
+\begin{figure}
+\begin{center}
+\includegraphics[scale=0.7]{../octave/norm_engine_model_plot}
+\includegraphics[scale=0.55]{../octave/norm_engine_plot_close}
+\end{center}
+\caption{Normalized engine model response.  After it stabilizes
+at unit step is applied to produce the normalized response.
+The lowest plot is a close up of the step response in the center
+plot.}
+\label{fig:normeng}
+\end{figure}
+% }}}
+
 % }}}
@@
 % }}}
+
+\clearpage
+\section{Conclusion}
 
@@
 
+\clearpage
+\lstinputlisting[
+	caption={Matlab source to plot the response of the normalized engine model.}
+	]{../octave/norm_engine_model_plot.m}
+
+\clearpage
+\lstinputlisting[
+	caption={Matlab source to plot the response of the normalized engine model
+	after step is applied.}
+	]{../octave/norm_engine_plot_close.m}
+
 \end{document}

### 2013-12-01 octave/norm_engine_model_plot.m

@@ -1,0 +1,52 @@
+%
+% norm_engine_model_plot.m
+%
+
+clear;
+
+ncyl = 8;
+
+engine_model;
+% eng_sys
+
+u0 = 600*1/1450.1;
+u1 = 1000*1/1450.1;
+
+% input signal
+u = [u0*ones(1,100) u1*ones(1,100)];
+
+% output response
+[y, t, x] = lsim(eng_sys, u);
+
+% An rpm of zero makes the rpmtime huge and
+% skews the results.
+% Remove these initial zeros.
+i = find(y~=0, 1, 'first');
+y = y(i:end);
+u = u(i:end);
+t = t(i:end);
+
+rt = rpmtime(y, 8);
+rt = cumsum(rt);
+
+% normalized output response
+yn = y*(1/400) - 3/2;
+
+figure;
+subplot(2,1,1);
+
+stairs(rt, u);
+grid on;
+axis ([rt(1), rt(end)]);
+title('Input Signal');
+ylabel('control');
+
+subplot(2,1,2);
+stairs(rt, yn);
+ylabel('Normalized RPM');
+grid on;
+axis ([rt(1), rt(end)]);
+title('Output Response');
+xlabel('time (sec)');
+
+print('norm_engine_model_plot.eps', '-deps');

### 2013-12-01 octave/norm_engine_plot_close.m
normalization corrections for engine model
Corrections to normalize the engine model have been found
which will accept a step input and produce a response near one.

@@ -1,0 +1,45 @@
+%
+% norm_engine_plot_close.m
+%
+
+clear;
+
+ncyl = 8;
+
+engine_model;
+% eng_sys
+
+u0 = 600*1/1450.1;
+u1 = 1000*1/1450.1;
+
+% input signal
+u = [u0*ones(1,100) u1*ones(1,100)];
+
+% output response
+[y, t, x] = lsim(eng_sys, u);
+
+% An rpm of zero makes the rpmtime huge and
+% skews the results.
+% Remove these initial zeros.
+i = find(y~=0, 1, 'first');
+y = y(i:end);
+u = u(i:end);
+t = t(i:end);
+
+rt = rpmtime(y, 8);
+rt = cumsum(rt);
+
+% normalized output response
+yn = y*(1/400) - 3/2;
+
+
+figure;
+
+stairs(rt(100:end), yn(100:end));
+grid on;
+axis([rt(100), rt(end) 0 1.5]);
+title('Normalized Engine Response After Step');
+xlabel('time (sec)');
+ylabel('normalized rpm');
+
+print('norm_engine_plot_close.eps', '-deps');

### 2013-12-01 octave/stable_rpm_plot.m

@@ -1,0 +1,31 @@
+%
+% stable_rpm_plot.m
+%
+
+clear;
+
+engine_model;
+
+% range of inputs to try
+us = linspace(0.2, 0.8, 7);
+
+y = zeros(0, length(us));
+for i = 1:length(us)
+	% response for a constant input
+	u = [us(i)*ones(1,100)];
+	[yp, t, x] = lsim(eng_sys, u);
+
+	% save the stable value
+	y(i) = yp(end);	
+endfor
+
+figure;
+
+plot(us, y, '-o');
+grid on;
+axis auto;
+title('Stable Output Rpm');
+xlabel('control input');
+ylabel('rpm');
+
+print('stable_rpm_plot.eps', '-deps');

### 2013-12-01 paper/main.tex
changed the title

@@
 
-\centerline{\Large \textbf{A Survey of Control Systems for Stabilizing}}
-\centerline{\Large \textbf{the Idle of a Spark Ignition Engine}}
+\centerline{\Large \textbf{Control System Design Applied to}}
+\centerline{\Large \textbf{Idle Stabilization of a Spark Ignition Engine}}
 \vspace{0.1in}
@@
 the problem of idle stabilization.
-Methods include: pole placement, PID, direct design, and
+Methods include: pole placement, direct design, and
 various state space designs.

### 2013-12-03 paper/main.tex
shifting/scaling unnecessary?
It appears that the shift and scaling of the model so
that it accommodates a step function may be unnecessary.
Systems designed from specs have no such requirement.

@@
 
+% prevent breaks in the middle of paragraphs
+\widowpenalties 1 10000
+\raggedbottom
+
 %
@@
 inputs are constant.
-This paper shows how control systems methods can be applied to
+This paper shows how control system methods can be applied to
 the problem of idle stabilization.
+Because the engine model is inherently discrete all of the methods
+used are also discrete.
 Methods include: pole placement, direct design, and
@@
 \footnotetext{The Matlab source of the RPM vs time plot is in Appendix \ref{app:rpmtime_plot_source}.}
-% }}}
-
-There is yet another complication with this model.
-Typical systems are in a normalized so that it is possible to
-achieve an output of one given a step input.
-With this engine model an output of 1 rpm is an impossibly slow value.
-And an input of zero or one rpm are similarly impossible.
-
-To resolve this issue the values will be scaled in to a normal range.
-A normal idle rpm range is from 600 to 1200 rpm.
-To scale these values so that zero equals 600 and one equals 1000
-Equation \ref{eq:scale} can be used.
-
-\begin{align}
-	y &= \frac{1}{400}x - \frac{3}{2} \label{eq:scale}
-\end{align}
-
-Figure \ref{fig:stable} shows the stable rpms for a given control input.
-Its behavior is described by Equation \ref{eq:stable}.
-
-\begin{align}
-	y &= 1450.1 x \label{eq:stable}
-\end{align}
-
-% {{{ fig:stable
-\begin{figure}[htbp!]
-\begin{center}
-\includegraphics[scale=0.6]{../octave/stable_rpm_plot}
-\end{center}
-\caption{Stable rpm reached for a given constant control input.}
-\label{fig:stable}
-\end{figure}
-% }}}
-
-With these normalization corrections applied the response can be
-seen in Figure \ref{fig:normeng}.
-To avoid the zero rpm case it is allowed to stabilize before the
-step is applied.
-The lowest plot shows the response after the step.
-It takes slightly over one second to stabilize, which is better than
-the inaccurate case from zero rpm, but it is still slow and can
-be improved.
-
-% {{{ fig:normeng
-\begin{figure}
-\begin{center}
-\includegraphics[scale=0.7]{../octave/norm_engine_model_plot}
-\includegraphics[scale=0.55]{../octave/norm_engine_plot_close}
-\end{center}
-\caption{Normalized engine model response.  After it stabilizes
-at unit step is applied to produce the normalized response.
-The lowest plot is a close up of the step response in the center
-plot.}
-\label{fig:normeng}
-\end{figure}
 % }}}

### 2013-12-03 octave/em_shift_plot2.m
created good examples of scaled/shifted model
The description of the engine response with a scaled/shifted
has been added.  The response should serve as a good basis
for building controllers and comparing the results.

@@ -1,0 +1,42 @@
+%
+% em_shift_plot2.m
+%
+% Engine model shifted plot close up.
+%
+
+clear;
+
+ncyl = 8;
+
+engine_model;
+% eng_sys
+
+% input signal (control)
+u_lo = 600/1450.1;
+u_hi = 1000/1450.1;
+n = 200;  % number of points
+u = [u_lo*ones(1,n/2) u_hi*ones(1,n/2)];
+
+% output response
+[y, t, x] = lsim(eng_sys, u);
+
+% An rpm of zero makes the rpmtime huge and skews the results.
+% Remove these initial zeros.
+i = find(y~=0, 1, 'first');
+y = y(i:end);
+u = u(i:end);
+t = t(i:end);
+
+rt = rpmtime(y, 8);
+rt = cumsum(rt);
+
+figure(1);
+
+stairs(rt((n/2):end), y((n/2):end));
+grid on;
+axis([rt(n/2), rt(end)]);
+title('Engine Response With Control From 0.41 to 0.69');
+xlabel('time (sec)');
+ylabel('rpm');
+
+print('em_shift_plot2.eps', '-deps');

### 2013-12-03 paper/main.tex
created good examples of scaled/shifted model
The description of the engine response with a scaled/shifted
has been added.  The response should serve as a good basis
for building controllers and comparing the results.

@@
 
+% {{{ fig:lem
 \begin{figure}[hbp!]
@@
 \end{figure}
+% }}}
 
@@
 \end{figure}
-\footnotetext{The Matlab source to produced these plots is in Appendix \ref{app:em_matlab}.}
+\footnotetext{The Matlab source for this and the other engine model plots
+are in Appendix \ref{app:em_matlab}.}
 % }}}
@@
 in Figure \ref{fig:rpmtime_plot}.
-It can be seen it takes over two seconds for the system to
-stabilize with no control.
-Clearly this is unacceptably slow and performance could be
-improved by using a controller.
+It can be seen that the rise time over $0.5$ second and
+that the overshoot is close to $40\%$.
+The stabilization time is also several seconds.
+Clearly its performance could be improved with the application
+of a controller.
 
@@
 \caption{Engine rpm response compared to time with a constant
-0.65 input control.\protect\footnotemark}
+0.65 input control.}
 \label{fig:rpmtime_plot}
 \end{figure}
-\footnotetext{The Matlab source of the RPM vs time plot is in Appendix \ref{app:rpmtime_plot_source}.}
+% }}}
+
+The response from zero rpm is not a realistic measurement of performance
+because zero rpm is impossibly slow.
+A typical idle rpm range is from 600 to 1000 rpm.
+
+Figure \ref{fig:em_shift_plot2} shows the response for a step input
+with control inputs chosen to reach 600 and 1000 rpm
+\footnote{The control input to stable rpm values are given in
+Appendix \ref{app:contrpm}}.
+This is a more realistic representation of typical operation.
+The rise time is near $0.2$ seconds and the overshoot is $15\%$.
+And the stabilization time is almost one second.
+Its behavior can still be improved with a controller.
+
+% {{{ fig:em_shift_plot2
+\begin{figure}[htbp!]
+\begin{center}
+\includegraphics[scale=0.6]{../octave/em_shift_plot2}
+\end{center}
+\caption{Response for a shifted and scaled step input with zero torque.}
+\label{fig:em_shift_plot2}
+\end{figure}
 % }}}
@@
 
+% {{{ Engine Model Matlab Source
 \clearpage
@@
 
+% XXX
 \clearpage
@@
 \lstinputlisting[
-	caption={Matlab source to plot the response of the normalized engine model.}
-	]{../octave/norm_engine_model_plot.m}
+	caption={Matlab source to plot the response of engine model
+		versus time for a shifted/scaled step input.}
+	]{../octave/em_shift_plot2.m}
+% }}}
+
+\clearpage
+\section{Control Response}
+\label{app:contrpm}
+
+For a given control input with zero torque the stable rpm values reached
+are shown in Figure \ref{fig:stable}.
+Equation \ref{eq:stable} describes its behavior.
+Listing \ref{lst:stable} shows how this was calculated.
+
+\begin{align}
+y &= 1450.1 x \label{eq:stable}
+\end{align}
+
+\begin{figure}[htbp!]
+\begin{center}
+\includegraphics[scale=0.5]{../octave/stable_rpm_plot}
+\end{center}
+\caption{Stable rpm reached for a given constant control input.}
+\label{fig:stable}
+\end{figure}
 
@@
 \lstinputlisting[
-	caption={Matlab source to plot the response of the normalized engine model
-	after step is applied.}
-	]{../octave/norm_engine_plot_close.m}
+	caption={Matlab source to find the stable rpm for a given
+		control input with zero torque.},
+	label={lst:stable}
+	]{../octave/stable_rpm_plot.m}
 

### 2013-12-03 octave/stable_rpm_plot.m

@@ -19,7 +19,7 @@ for i = 1:length(us)
 	y(i) = yp(end);	
 endfor
 
-figure;
+figure(1);
 
 plot(us, y, '-o');
 grid on;

### 2013-12-03 paper/main.tex
cleaned up diagram and some text, minor

@@
 The configuration is a modern 4.6L V-8 gas engine.
-A single control input representing the idle air bypass valve
-
-To simplify analysis the linearized model is used as shown in
+To simplify this analysis the linearized model was used as shown in
 Figure \ref{fig:lem}.
@@
 
-\begin{tikzpicture}[>=triangle 60, scale=0.7,
+\begin{tikzpicture}[>=triangle 60, scale=0.6,
 	node distance=14mm, auto]
@@
 
-	\draw [->] (-20mm,0) node[above]{$\Delta \text{control}$} -- (gain1.west);
+	\draw [->] (-30mm,0) node[above,xshift=2mm,yshift=1mm]{$\Delta \text{control}$}
+				-- (gain1.west);
 	\draw [->] (gain1) -- (sum1);
@@
 	\draw [->] (sum2) -- (gain3);
-	\draw [->] (gain3.east) -- ++(20mm,0) node[above]{$\Delta \text{rpm}$};
+	\draw [->] (gain3.east) -- ++(30mm,0) node[above,xshift=-5mm,yshift=1mm]{$\Delta \text{rpm}$};
 	\draw [->] ($(gain3.east) + (10mm,0) $) |- (gain4.east);
@@
 which is at a minimum near zero and at a maximum near unity.
-Often the duty cycle range is in a range from 0\% to 100\% which
-corresponds to 0 to 1 (unity).
 
@@
 this model cannot be used directly with typical control systems
-which use steady state values.
-It is possible convert these deltas to steady state equivalents.
+which expect steady state values.
+It is possible to convert these deltas to steady state equivalents.
 Figure \ref{fig:sdtf} shows the transfer function to convert steady
-state values to delta values.
+state values to delta values
+\footnote{The derivation of the steady state to delta conversion
+is given in Appendix \ref{app:cdelta}}.
 Figure \ref{fig:dstf} shows the transfer function to delta values
-to steady state values.
+to steady state values
+\footnote{The derivation of the delta to steady state conversion
+is given in Appendix \ref{app:dsderiv}.}.
 
@@
 \begin{figure}[!htbp]
-
 \begin{center}
@@
 \end{center}
+\caption[xyz]{The $Z$ transform used to accumulate the input and convert
+a steady state input to delta output.}
+\label{fig:sdtf}
+\end{figure}
 
-\caption{The $Z$ transform used to accumulate the input and convert
-a steady state input to delta output.
-Its derivation is given in Appendix \ref{app:cdelta}.}
-\label{fig:sdtf}
-
-\end{figure}
 % }}}
@@
 \caption{The $Z$ transform used to convert delta input to a steady
-state output.
-Its derivation is given in Appendix \ref{app:dsderiv}.}
+state output.}
 \label{fig:dstf}
@@
 
-% TODO - reduced steady state system
-
-In order to build a controller the model needs to be simplified
-but the presence of two inputs complicates matters.
+In order to construct a controller the model (Figure \ref{fig:lem})
+needs to be simplified in to a transfer function.
+However the prescence of two inputs, control and torque, complicates matters.
 To resolve this issue the torque can be set zero.
@@
 
-First taking the parts from the engine model (Figure \ref{fig:lem}).
+To simply first gather the parts from the engine model (Figure \ref{fig:lem}).
 
@@
 \begin{samepage}
-Then substituting them in to Equation \ref{eq:dd} and simplifying
-results in Equation \ref{eq:lems}.
+Then substitute them in to Equation \ref{eq:dd} and simplify.
+The result is Equation \ref{eq:lems}.
 \begin{align}
@@
 It is still necessary to address the delta inputs and outputs.
-This can be resolved by placing the conversion transforms,
-from Figure \ref{fig:sdtf} and \ref{fig:dstf}, on either
-end of the engine model transfer function as shown in Figure \ref{fig:nodelta}.
+This can be resolved by placing the conversion transforms
+(from Figure \ref{fig:sdtf} and \ref{fig:dstf}) on either
+end of the engine model as shown in Figure \ref{fig:nodelta}.
 A beneficial side effect becomes apparent in this form.
@@
 with the transform that converts from a delta to steady state
-\footnote{In the more general case, when torque is non zero,
-the transforms would not cancel.}.
+\footnote{In the more general case, when torque is not zero,
+the transforms wouldn't cancel.}.
 Therefore the final transform is still Equation \ref{eq:lems}.
@@
 Figure \ref{fig:em_plot}.
+The Matlab source code for this plot and others are
+in Appendix \ref{app:em_matlab}.
 \end{samepage}
@@
 Here the control is constant at 0.65 which resulted in a stable
-rpm near 900.\protect\footnotemark}
+rpm near 900.}
 \label{fig:em_plot}
 \end{figure}
-\footnotetext{The Matlab source for this and the other engine model plots
-are in Appendix \ref{app:em_matlab}.}
 % }}}
@@
 It is not apparent from Figure \ref{fig:em_plot} exactly how long
-this process takes.
-The $x$ axis represents ignition events, not time.
-The rpm can be converted to time per ignition event by examining
+this process takes because the $x$ axis represents ignition events
+instead of time.
+However the rpm can be converted to time per ignition event by examining
 the units as shown in Equation \ref{eq:rpmtime}.
@@
 
-Applying this equation results in the response shown
+Applying Equation \ref{eq:rpmtime} results in the response shown
 in Figure \ref{fig:rpmtime_plot}.
-It can be seen that the rise time over $0.5$ second and
+It can be seen that the rise time is over $0.5$ second and
 that the overshoot is close to $40\%$.
@@
 A typical idle rpm range is from 600 to 1000 rpm.
-
 Figure \ref{fig:em_shift_plot2} shows the response for a step input

### 2013-12-03 octave/norm_engine_model_plot.m

@@ -1,52 +1,0 @@
-%
-% norm_engine_model_plot.m
-%
-
-clear;
-
-ncyl = 8;
-
-engine_model;
-% eng_sys
-
-u0 = 600*1/1450.1;
-u1 = 1000*1/1450.1;
-
-% input signal
-u = [u0*ones(1,100) u1*ones(1,100)];
-
-% output response
-[y, t, x] = lsim(eng_sys, u);
-
-% An rpm of zero makes the rpmtime huge and
-% skews the results.
-% Remove these initial zeros.
-i = find(y~=0, 1, 'first');
-y = y(i:end);
-u = u(i:end);
-t = t(i:end);
-
-rt = rpmtime(y, 8);
-rt = cumsum(rt);
-
-% normalized output response
-yn = y*(1/400) - 3/2;
-
-figure;
-subplot(2,1,1);
-
-stairs(rt, u);
-grid on;
-axis ([rt(1), rt(end)]);
-title('Input Signal');
-ylabel('control');
-
-subplot(2,1,2);
-stairs(rt, yn);
-ylabel('Normalized RPM');
-grid on;
-axis ([rt(1), rt(end)]);
-title('Output Response');
-xlabel('time (sec)');
-
-print('norm_engine_model_plot.eps', '-deps');

### 2013-12-05 octave/cd1_init.m
Direct Design working, major cleanup of Octave files
The first model has been built using Direct Design.
And it is a significant improvement compared to no control.

The Octave files used for generating the plots have been
cleaned up and their names made more consistent.

@@ -1,16 +1,0 @@
-% 
-% cd1_init.m
-%
-% Convert a steady state input to a delta output.
-%
-% Full System.
-%
-
-T = 1;  % time step
-
-D1 = tf([1], [1 0], T, 'inname', 'y1', 'outname', 'v1');
-D2 = tf([1], [1 0], T, 'inname', 'r1', 'outname', 'q1');
-sum1 = sumblk('y1 = u1 - r1');
-sum2 = sumblk('r1 = v1 + q1');
-sys = connect(D1, D2, sum1, sum2, 'u1', 'y1');
-

### 2013-12-05 octave/cd2_init.m

@@ -1,11 +1,0 @@
-% 
-% cd2_init.m
-%
-% Convert a steady state input to a delta output.
-%
-% Simplified System.
-%
-
-T = 1;  % time step
-
-sys = tf([1 -1], [1 0], T, 'inname', 'y1', 'outname', 'u1');

### 2013-12-05 octave/cd_plot1.m

@@ -1,0 +1,31 @@
+%
+% cd_plot1.m
+%
+
+clear;
+
+T = 1;  % time step
+
+D1 = tf([1], [1 0], T, 'inname', 'y1', 'outname', 'v1');
+D2 = tf([1], [1 0], T, 'inname', 'r1', 'outname', 'q1');
+sum1 = sumblk('y1 = u1 - r1');
+sum2 = sumblk('r1 = v1 + q1');
+sys = connect(D1, D2, sum1, sum2, 'u1', 'y1');
+
+u = [1 1 1 -3 -3 2 5 0];
+t = 0:(size(u,2)-1);  % start at zero
+
+figure;
+subplot(2,1,1);
+stairs(t,u);
+grid on;
+axis auto;
+title('Input Signal');
+ylabel('u');
+
+subplot(2,1,2);
+lsim(sys, u);
+grid on;
+axis auto;
+
+print('cd_plot1.eps', '-color', '-deps2');

### 2013-12-05 octave/cd_plot2.m

@@ -1,12 +1,14 @@
 %
-% dc1_plot.m
+% cd2_plot.m
 %
 
 clear;
 
-dc1_init;
+T = 1;  % time step
 
-u = [1 0 0 -4 0 5 3 3];
+sys = tf([1 -1], [1 0], T, 'inname', 'y1', 'outname', 'u1');
+
+u = [1 1 1 -3 -3 2 5 0];
 t = 0:(size(u,2)-1);  % start at zero
 
 figure;
@@ -22,4 +24,4 @@ lsim(sys, u);
 grid on;
 axis auto;
 
-print('dc1_plot.eps', '-deps');
+print('cd_plot2.eps', '-color', '-deps2');

### 2013-12-05 octave/dd_plot2.m

@@ -1,0 +1,72 @@
+%
+% dd_plot1.m
+%
+% Plot of solution using direct design compared to no control.
+%
+
+clear;
+
+T = 1;
+ncyl = 8;
+
+% input signal
+u_lo = 600;		% rpm before step
+u_hi = 1000;	% rpm after step
+% different systems take longer to stabilize (time lo)
+n1_lo = 30;		% nth tick to step from lo to hi
+n2_lo = 100;
+n_hi = 30;		% ticks after step
+u1 = [u_lo*ones(1,n1_lo) u_hi*ones(1,n_hi)];
+% convert to control values (without control)
+u2 = [(u_lo/1450.1)*ones(1,n2_lo) (u_hi/1450.1)*ones(1,n_hi)];
+
+% Build a controller, and the resulting system.
+Gz1 = engine_model(T);
+% These roots are guess, taken from some other example.
+roots = [(0.6 + 0.4i) (0.6 - 0.4i) 0 0 0];
+order = 3;
+limit = 1;
+[Hz1, Dz1, K1] = direct_design(Gz1, roots, order, limit);
+
+% System with no control.
+Gz2 = engine_model(T);
+
+% output response
+[y1, t1, x, i1] = lsim1(Hz1, u1);
+[y2, t2, x, i2] = lsim1(Gz2, u2);  % no control
+% adjust the step point for any removed from the beginning
+i = max([i1 i2]);
+
+% convert rpm to time
+rt1 = rpmtime(y1, ncyl);
+rt2 = rpmtime(y2, ncyl);
+
+% remove data before step is applied
+p1_start = (length(rt1) - n_hi) + 1;
+rt1 = rt1(p1_start:end);
+y1 = y1(p1_start:end);
+
+p2_start = (length(rt2) - n_hi) + 1;
+rt2 = rt2(p2_start:end);
+y2 = y2(p2_start:end);
+
+% reset times to start at zero
+rt1 = zerotime(rt1);
+rt2 = zerotime(rt2);
+
+% plot
+
+clf;
+figure(1);
+
+[xs1, ys1] = stairs(rt1, y1);
+[xs2, ys2] = stairs(rt2, y2);
+plot(xs2, ys2, xs1, ys1);
+
+grid on;
+title(sprintf('Output Response, step: %d - %d rpm', u_lo, u_hi));
+ylabel('rpm');
+xlabel('time (sec)');
+legend('no control', 'direct design');
+
+print('dd_plot2.eps', '-color', '-deps2');

### 2013-12-05 octave/em_plot4.m

@@ -1,0 +1,39 @@
+%
+% em_plot4.m
+%
+% Engine model with no control.
+% Allowed to stablize before step is applied.
+%
+
+T = 1;
+
+Gz = engine_model(T);
+
+% input signal (control)
+u_lo = 600/1450.1;
+u_hi = 1000/1450.1;
+n = 200;  % number of points
+u = [u_lo*ones(1,n/2) u_hi*ones(1,n/2)];
+
+% output response
+[y, t, x,i] = lsim1(Gz, u);
+
+rt = rpmtime(y, 8);
+
+% remove data before the step
+rt = rt((n/2):end);
+y = y((n/2):end);
+
+% reset start time to zero
+rt = zerotime(rt);
+
+figure(1);
+
+stairs(rt, y);
+grid on;
+axis([rt(1), rt(end)]);
+title('Engine Response With Control From 0.41 to 0.69');
+xlabel('time (sec)');
+ylabel('rpm');
+
+print('em_plot4.eps', '-color', '-deps2');

### 2013-12-05 octave/engine_model.m

@@ -1,10 +1,10 @@
-% 
-% engine_model.m
+function [sys] = engine_model(T);
+%  ENGINE_MODEL
 %
 % Linear engine model with zero torque.
 %
 
-T = 1;  % time step
+%T = 1;  % time step
 
 K = tf([1.699], [1], T);
 D1= tf([8.5683], [1 -0.9025 0], T);
@@ -14,4 +14,6 @@ D = tf([0.00093], [1], T);
 
 H = K*G/(1 + D*G);
 
-eng_sys = minreal(H);  % cancel common roots
+sys = minreal(H);  % cancel common roots
+
+endfunction

### 2013-12-05 octave/engine_model_plot.m

@@ -1,28 +1,0 @@
-%
-% engine_model_plot.m
-%
-
-clear;
-
-engine_model;
-
-u = [0.65*ones(1,100)];
-[y, t, x] = lsim(eng_sys, u);
-
-figure;
-subplot(2,1,1);
-
-stairs(t, u);
-grid on;
-axis auto;
-title('Input Signal');
-ylabel('control');
-
-subplot(2,1,2);
-stairs(t, y);
-grid on;
-axis auto;
-title('Output Response');
-ylabel('rpm');
-
-print('engine_model_plot.eps', '-deps');

### 2013-12-05 octave/lsim1.m

@@ -1,0 +1,24 @@
+function [y, t, x, i] = lsim1(Hz, u);
+% LSIM1 - Just like lsim excpt small values are removed.
+%
+% "small" means less than or equal to 10.
+%
+
+[y, t, x] = lsim(Hz, u);
+
+% Find the first valid value
+for i = 1:length(y) 
+	x = abs(y(i));
+	if (x > 10)
+		break;
+	end
+endfor
+
+% adjust for removed values
+y = y(i:end);
+u = u(i:end);
+t = t(i:end);
+
+i = i - 1;
+
+endfunction

### 2013-12-05 paper/main.tex
Direct Design working, major cleanup of Octave files
The first model has been built using Direct Design.
And it is a significant improvement compared to no control.

The Octave files used for generating the plots have been
cleaned up and their names made more consistent.

@@
 \begin{center}
-\includegraphics[scale=0.7]{../octave/engine_model_plot}
+\includegraphics[scale=0.7]{../octave/em_plot1}
 \end{center}
@@
 \begin{center}
-\includegraphics[scale=0.7]{../octave/engine_model_rpmtime_plot}
+\includegraphics[scale=0.7]{../octave/em_plot2}
 \end{center}
@@
 A typical idle rpm range is from 600 to 1000 rpm.
-Figure \ref{fig:em_shift_plot2} shows the response for a step input
+Figure \ref{fig:em_plot4} shows the response for a step input
 with control inputs chosen to reach 600 and 1000 rpm
@@
 
-% {{{ fig:em_shift_plot2
+% {{{ fig:em_plot4
 \begin{figure}[htbp!]
 \begin{center}
-\includegraphics[scale=0.6]{../octave/em_shift_plot2}
+\includegraphics[scale=0.6]{../octave/em_plot4}
 \end{center}
 \caption{Response for a shifted and scaled step input with zero torque.}
-\label{fig:em_shift_plot2}
+\label{fig:em_plot4}
 \end{figure}
@@
 
-% {{{ Pole Placement
-\section{Pole Placement}
+% {{{ Direct Design
+\section{Direct Design}
+
+% {{{ fig:dd_plot2
+\begin{figure}[htbp!]
+\begin{center}
+\includegraphics[scale=0.7]{../octave/dd_plot2}
+\end{center}
+\caption{Output response of controller built using Direct Design
+compared to no control.  For no control this is the same plot
+as in Figure \ref{fig:em_plot4} which stabilizes after a second.}
+\label{fig:dd_plot2}
+\end{figure}
+% }}}
 
@@
 
-Figure \ref{fig:cd1_plot} shows the response of this system given
+Figure \ref{fig:cd_plot1} shows the response of this system given
 an arbitrary input.
@@
 \begin{figure}[htbp!]
-\includegraphics[scale=0.7]{../octave/cd1_plot}
+\includegraphics[scale=0.7]{../octave/cd_plot1}
 \caption{Response of full steady state input to delta output system
@@
 the output response ($y$).
-The Matlab source code is given in Listing \ref{lst:cd1_init}
-and \ref{lst:cd1_plot}.
+The Matlab source code is given in Listing \ref{lst:cd_plot1}.
 }
-\label{fig:cd1_plot}
+\label{fig:cd_plot1}
 \end{figure}
@@
 
-It can be seen in Figure \ref{fig:cd2_plot} that the simplified system
-behaves identically to the previous system (Figure \ref{fig:cd1_plot}).
+It can be seen in Figure \ref{fig:cd_plot2} that the simplified system
+behaves identically to the previous system (Figure \ref{fig:cd_plot1}).
 
 \begin{figure}[htbp!]
-\includegraphics[scale=0.7]{../octave/cd2_plot}
+\includegraphics[scale=0.7]{../octave/cd_plot2}
 \caption{Response of simplified steady state input to delta output system
@@
 the output response ($y$).
-Response is identical to the full system in Figure \ref{fig:cd1_plot}
+Response is identical to the full system in Figure \ref{fig:cd_plot1}
 as expected.
-The Matlab source code is given in Listing \ref{lst:cd2_init}
-and \ref{lst:cd2_plot}.
+The Matlab source code is given in Listing \ref{lst:cd_plot2}.
 }
-\label{fig:cd2_plot}
+\label{fig:cd_plot2}
 \end{figure}
@@
 
-\lstinputlisting[caption={Matlab code to initialize the full
-steady state to delta system.},
-	label=lst:cd1_init
-	]{../octave/cd1_init.m}
-
 \lstinputlisting[caption={Matlab code to plot the full steady state
 to delta system.},
-	label=lst:cd1_plot
-	]{../octave/cd1_plot.m}
+	label=lst:cd_plot1
+	]{../octave/cd_plot1.m}
 
 \clearpage
-\lstinputlisting[caption={Matlab code to initialize the simplified
-steady state to delta system.},
-	label=lst:cd2_init
-	]{../octave/cd2_init.m}
-
 \lstinputlisting[caption={Matlab code to plot the simplified steady state
 to delta system.},
-	label=lst:cd2_plot
-	]{../octave/cd2_plot.m}
+	label=lst:cd_plot2
+	]{../octave/cd_plot2.m}
 
@@
 
-Figure \ref{fig:dc2_plot} shows the response of this system given
+Figure \ref{fig:dc_plot2} shows the response of this system given
 an arbitrary input
@@
 
-% {{{ fig:dc2_plot
+% {{{ fig:dc_plot2
 \begin{figure}[htbp!]
-\includegraphics[scale=0.7]{../octave/dc2_plot}
+\includegraphics[scale=0.7]{../octave/dc_plot2}
 \caption{Response of delta input to steady state output system
@@
 the output response ($y$).
-The Matlab source code is given in Listing \ref{lst:dc2_init}
-and \ref{lst:dc2_plot}.
+The Matlab source code is given in Listing \ref{lst:dc_plot2}.
 }
-\label{fig:dc2_plot}
+\label{fig:dc_plot2}
 \end{figure}
@@
 
-%\lstinputlisting[caption={Matlab code to initialize the full
-%steady state to delta system.},
-%	label=lst:dc1_init
-%	]{../octave/dc1_init.m}
-%
-%\lstinputlisting[caption={Matlab code to plot the full steady state
-%to delta system.},
-%	label=lst:dc1_plot
-%	]{../octave/dc1_plot.m}
-
-\lstinputlisting[caption={Matlab code to initialize the simplified
-steady state to delta system.},
-	label=lst:dc2_init
-	]{../octave/dc2_init.m}
-
 \lstinputlisting[caption={Matlab code to plot the simplified steady state
 to delta system.},
-	label=lst:dc2_plot
-	]{../octave/dc2_plot.m}
+	label=lst:dc_plot2
+	]{../octave/dc_plot2.m}
 
@@
 	caption={Matlab source to calculate the engine model system with
-		zero torque.}
+		zero torque.  This function is called by other scripts that
+	use this model.}
 	]{../octave/engine_model.m}
@@
 	caption={Matlab source to plot the response of engine model
-		with no control.}
-	]{../octave/engine_model_plot.m}
+		with no control. The $x$ axis is in ticks.}
+	]{../octave/em_plot1.m}
 
@@
 	caption={Matlab source to plot the response of engine model
-		versus time.}
-	]{../octave/engine_model_rpmtime_plot.m}
+	with no control.  The $x$ axis is in time.}
+	]{../octave/em_plot2.m}
 
@@
 	caption={Matlab source to plot the response of engine model
-		versus time for a shifted/scaled step input.}
-	]{../octave/em_shift_plot2.m}
+		versus time for a typical 600 to 1000 rpm range.}
+	]{../octave/em_plot4.m}
 % }}}
 
+% {{{ Direct Design Matlab Source
+\clearpage
+\section{Direct Design Matlab Source}
+\label{app:dd_matlab}
+
+\lstinputlisting[
+	caption={Matlab source used to plot the controller built using
+	Direct Design compared to no control.}
+	]{../octave/dd_plot2.m}
+
+% }}}
+
+% {{{ Control Response
 \clearpage
@@
 \begin{center}
-\includegraphics[scale=0.5]{../octave/stable_rpm_plot}
+\includegraphics[scale=0.5]{../octave/em_plot3}
 \end{center}
@@
 	caption={Matlab source to find the stable rpm for a given
-		control input with zero torque.},
+		control input.},
 	label={lst:stable}
-	]{../octave/stable_rpm_plot.m}
+	]{../octave/em_plot3.m}
+% }}}
 

### 2013-12-05 octave/cd1_plot.m

@@ -1,25 +1,0 @@
-%
-% cd1_plot.m
-%
-
-clear;
-
-cd1_init;
-
-u = [1 1 1 -3 -3 2 5 0];
-t = 0:(size(u,2)-1);  % start at zero
-
-figure;
-subplot(2,1,1);
-stairs(t,u);
-grid on;
-axis auto;
-title('Input Signal');
-ylabel('u');
-
-subplot(2,1,2);
-lsim(sys, u);
-grid on;
-axis auto;
-
-print('cd1_plot.eps', '-deps');

### 2013-12-05 octave/cd2_plot.m

@@ -1,25 +1,0 @@
-%
-% cd2_plot.m
-%
-
-clear;
-
-cd2_init;
-
-u = [1 1 1 -3 -3 2 5 0];
-t = 0:(size(u,2)-1);  % start at zero
-
-figure;
-subplot(2,1,1);
-stairs(t,u);
-grid on;
-axis auto;
-title('Input Signal');
-ylabel('u');
-
-subplot(2,1,2);
-lsim(sys, u);
-grid on;
-axis auto;
-
-print('cd2_plot.eps', '-deps');

### 2013-12-05 octave/dc1_init.m

@@ -1,13 +1,0 @@
-% 
-% dc1_init.m
-%
-% Convert a delta input to a steady state output.
-%
-% Full System.
-%
-
-T = 1;  % time step
-
-D1 = tf([1], [1 0], T, 'inname', 'y1', 'outname', 'r1');
-S1 = sumblk('y1 = u1 + r1');
-sys = connect(D1, S1, 'u1', 'y1');

### 2013-12-05 octave/dc2_init.m

@@ -1,11 +1,0 @@
-% 
-% dc1_init.m
-%
-% Convert a delta input to a steady state output.
-%
-% Simplified System.
-%
-
-T = 1;  % time step
-
-sys = tf([1 0], [1 -1], T, 'inname', 'u1', 'outname', 'y1');

### 2013-12-05 octave/dc_plot1.m

@@ -1,0 +1,31 @@
+%
+% dc_plot.m
+%
+% Digital to steady state transfer function.
+%
+
+clear;
+
+T = 1;  % time step
+
+D1 = tf([1], [1 0], T, 'inname', 'y1', 'outname', 'r1');
+S1 = sumblk('y1 = u1 + r1');
+sys = connect(D1, S1, 'u1', 'y1');
+
+u = [1 0 0 -4 0 5 3 3];
+t = 0:(size(u,2)-1);  % start at zero
+
+figure;
+subplot(2,1,1);
+stairs(t,u);
+grid on;
+axis auto;
+title('Input Signal');
+ylabel('u');
+
+subplot(2,1,2);
+lsim(sys, u);
+grid on;
+axis auto;
+
+print('dc_plot1.eps', '-color', '-deps2');

### 2013-12-05 octave/dc_plot2.m

@@ -1,10 +1,12 @@
 %
-% dc2_plot.m
+% dc_plot2.m
 %
 
 clear;
 
-dc2_init;
+T = 1;  % time step
+
+sys = tf([1 0], [1 -1], T, 'inname', 'u1', 'outname', 'y1');
 
 u = [1 0 0 -4 0 5 3 3];
 t = 0:(size(u,2)-1);  % start at zero
@@ -24,4 +26,4 @@ grid on;
 axis auto;
 axis([0 7 -4 6]);
 
-print('dc2_plot.eps', '-deps');
+print('dc_plot2.eps', '-color', '-deps2');

### 2013-12-05 octave/dd_plot1.m

@@ -1,0 +1,49 @@
+%
+% dd_plot1.m
+%
+% Direct Design controller close up.
+%
+
+clear;
+
+T = 1;
+
+% input signal
+u_lo = 600;		% rpm before step
+u_hi = 1000;	% rpm after step
+n_lo = 30;		% nth tick to step from lo to hi
+n_hi = 30;		% ticks after step
+u = [u_lo*ones(1,n_lo) u_hi*ones(1,n_hi)];
+
+% build a controller, and the resulting system
+Gz = engine_model(T);
+roots = [(0.6 + 0.4i) (0.6 - 0.4i) 0 0 0];
+order = 3;
+limit = 1;
+[Hz, Dz, K] = direct_design(Gz, roots, order, limit);
+
+% output response
+[y, t, x, i] = lsim1(Hz, u);
+% adjust the step point for any removed from the beginning
+n_lo -= i;
+
+% convert rpm to time
+rt = rpmtime(y, 8);
+
+clf;
+figure(1);
+
+% reduce data to what we are interested in
+%p_start = 1;  	% all data
+p_start = n_lo + 1;	% at step
+rt = rt(p_start:end);
+y = y(p_start:end);
+
+stairs(rt, y);
+axis([rt(1), rt(end)]);
+grid on;
+title(sprintf('Output Response, step: %d - %d rpm', u_lo, u_hi));
+ylabel('rpm');
+xlabel('time (sec)');
+
+print('dd_plot1.eps', '-color', '-deps2');

### 2013-12-05 octave/direct_design.m

@@ -1,0 +1,44 @@
+function [Hz, Dz, K] = direct_design(Gz, roots, order, limit=1);
+% DIRECT_DESIGN - Construct a controller using Direct Design.
+%
+% 'Gz' is some plant to be controlled.
+% 'order' is the order of the system need for the Sylvester Matrix.
+% 'roots' are the designer provided roots to achieve in the system (Hz).
+%
+% The entire system is returned (Hz) along with the controller (Dz)
+% and the constant (K) that were used.
+%
+
+% Designer provided specifications.
+%
+n = order;
+D = poly(roots);
+if (isrow(D))
+	D = transpose(D);
+end
+% Reverse D so Alpha and Beta are in the correct order.
+D = flipud(D);
+
+[Bz, Az, T] = tfdata(Gz, 'v');
+
+E = sylvester(Az, Bz);
+
+%M = E^-1*D;
+M = E\D;
+
+% Alpha = a0*z + a1
+% Beta = b0*z + b1
+Alpha = fliplr(transpose(M(1:n)));
+Beta = fliplr(transpose(M((n+1):end)));
+
+Dz = tf(Beta, Alpha, T);
+
+% To find K, the limit should go to 1
+% for a step input.
+% (refer to the notes for a better description)
+K = sum(D)/sum(conv(Bz, Alpha));
+% XXX
+
+Hz = K*Gz/(1 + Dz*Gz);
+
+endfunction

### 2013-12-05 octave/em_plot1.m

@@ -1,0 +1,33 @@
+%
+% em_plot.m
+%
+% Plot of engine model with zero torque and no
+% control.  x axis is in ticks (not time).
+%
+
+clear;
+
+T = 1;  % time step
+Gz = engine_model(T);
+
+u = [0.65*ones(1,100)];
+[y, t, x] = lsim(Gz, u);
+
+figure;
+subplot(2,1,1);
+
+stairs(t, u);
+grid on;
+axis([t(1) t(end)]);
+title('Input Signal');
+ylabel('control');
+
+subplot(2,1,2);
+stairs(t, y);
+grid on;
+axis([t(1) t(end)]);
+title('Output Response');
+ylabel('rpm');
+xlabel('ticks');
+
+print('em_plot1.eps', '-color', '-deps2');

### 2013-12-05 octave/em_plot2.m

@@ -1,15 +1,18 @@
 %
-% engine_model_rpmtime_plot.m
+% em_plot2.m
+%
+% Engine model with no control and x axis in time.
 %
 
 clear;
 
 ncyl = 8;
 
-engine_model;
+T = 1;  % time step
+Gz = engine_model(T);
 
 u = [0.65*ones(1,100)];
-[rpm, t, x] = lsim(eng_sys, u);
+[rpm, t, x] = lsim(Gz, u);
 
 % An rpm of zero makes the rpmtime huge and
 % skews the results.
@@ -23,22 +26,21 @@ t = t(i:end);
 % The cummulative sum is then the time it takes
 % to get to that particular point.
 rt = rpmtime(rpm, 8);
-rt = cumsum(rt);
 
 figure;
 subplot(2,1,1);
 stairs(rt, u);
 grid on;
-axis auto;
+axis([rt(1), rt(end)]);
 title('Input Signal');
 ylabel('control');
 
 subplot(2,1,2);
 stairs(rt, rpm);
 grid on;
-axis auto;
+axis([rt(1), rt(end)]);
 title('Output Response');
 ylabel('rpm');
 xlabel('time (sec)');
 
-print('engine_model_rpmtime_plot.eps', '-deps');
+print('em_plot2.eps', '-color', '-deps2');

### 2013-12-05 octave/em_plot3.m

@@ -1,10 +1,13 @@
 %
-% stable_rpm_plot.m
+% em_plot3.m
+%
+% Plot of stable output rpm's for a given control input.
 %
 
 clear;
 
-engine_model;
+T = 1;
+Gz = engine_model(T);
 
 % range of inputs to try
 us = linspace(0.2, 0.8, 7);
@@ -13,7 +16,7 @@ y = zeros(0, length(us));
 for i = 1:length(us)
 	% response for a constant input
 	u = [us(i)*ones(1,100)];
-	[yp, t, x] = lsim(eng_sys, u);
+	[yp, t, x] = lsim(Gz, u);
 
 	% save the stable value
 	y(i) = yp(end);	
@@ -28,4 +31,4 @@ title('Stable Output Rpm');
 xlabel('control input');
 ylabel('rpm');
 
-print('stable_rpm_plot.eps', '-deps');
+print('em_plot3.eps', '-color', '-deps2');

### 2013-12-05 octave/em_shift_plot2.m

@@ -1,42 +1,0 @@
-%
-% em_shift_plot2.m
-%
-% Engine model shifted plot close up.
-%
-
-clear;
-
-ncyl = 8;
-
-engine_model;
-% eng_sys
-
-% input signal (control)
-u_lo = 600/1450.1;
-u_hi = 1000/1450.1;
-n = 200;  % number of points
-u = [u_lo*ones(1,n/2) u_hi*ones(1,n/2)];
-
-% output response
-[y, t, x] = lsim(eng_sys, u);
-
-% An rpm of zero makes the rpmtime huge and skews the results.
-% Remove these initial zeros.
-i = find(y~=0, 1, 'first');
-y = y(i:end);
-u = u(i:end);
-t = t(i:end);
-
-rt = rpmtime(y, 8);
-rt = cumsum(rt);
-
-figure(1);
-
-stairs(rt((n/2):end), y((n/2):end));
-grid on;
-axis([rt(n/2), rt(end)]);
-title('Engine Response With Control From 0.41 to 0.69');
-xlabel('time (sec)');
-ylabel('rpm');
-
-print('em_shift_plot2.eps', '-deps');

### 2013-12-05 octave/norm_engine_plot_close.m

@@ -1,45 +1,0 @@
-%
-% norm_engine_plot_close.m
-%
-
-clear;
-
-ncyl = 8;
-
-engine_model;
-% eng_sys
-
-u0 = 600*1/1450.1;
-u1 = 1000*1/1450.1;
-
-% input signal
-u = [u0*ones(1,100) u1*ones(1,100)];
-
-% output response
-[y, t, x] = lsim(eng_sys, u);
-
-% An rpm of zero makes the rpmtime huge and
-% skews the results.
-% Remove these initial zeros.
-i = find(y~=0, 1, 'first');
-y = y(i:end);
-u = u(i:end);
-t = t(i:end);
-
-rt = rpmtime(y, 8);
-rt = cumsum(rt);
-
-% normalized output response
-yn = y*(1/400) - 3/2;
-
-
-figure;
-
-stairs(rt(100:end), yn(100:end));
-grid on;
-axis([rt(100), rt(end) 0 1.5]);
-title('Normalized Engine Response After Step');
-xlabel('time (sec)');
-ylabel('normalized rpm');
-
-print('norm_engine_plot_close.eps', '-deps');

### 2013-12-05 octave/rpmtime.m

@@ -1,25 +1,28 @@
-function [tign] = rpmtime(rpm, ncyl)
-% RPMTIME Calculate the time (in seconds) between ignition
-%  events at a given rpm.
+function [ts] = rpmtime(rpm, ncyl)
+% RPMTIME Calculate the cummulative time (in seconds) between
+%  ignition events at a given rpm.
 %
-%  rpmtime(rpm, ncyl);
-%  rpmtime(400, 8);
+%  [ts] = rpmtime(rpm, ncyl);
+%  [ts  = rpmtime(400, 8);
 %
 
-	%ncyl;				% number of cylinders
-	%rpm;				% rpm
-	ign_rev = ncyl/2;	% igntion points per revolution
-	min_sec = 1/60;		% 1 minute per 60 seconds
-	%tign;				% seconds per ignition event
+%ncyl;				% number of cylinders
+%rpm;				% rpm
+ign_rev = ncyl/2;	% igntion points per revolution
+min_sec = 1/60;		% 1 minute per 60 seconds
+%ts	;				% seconds per ignition event
 
-	tign = zeros(0,length(rpm));
-	for i = 1:length(rpm)
-		x = (rpm(i)*min_sec*ign_rev);
-		if (0 == x)
-			tign(i) = inf;
-		else
-			tign(i) = 1/x;
-		endif
-	endfor
+ts = zeros(0,length(rpm));  % initial empty vector
+for i = 1:length(rpm)
+	x = (rpm(i)*min_sec*ign_rev);
+
+	if (0 == x)
+		ts(i) = inf;
+	else
+		ts(i) = 1/x;
+	endif
+endfor
+
+ts = cumsum(ts);
 
 endfunction

### 2013-12-05 octave/sylvester.m

@@ -1,0 +1,33 @@
+function [E] = sylvester(A, B)
+% SYLVESTER Construct a Sylvester matrix of the two vectors.
+%
+% Given two vectors, the largest order is determined.
+% Then a Sylvester matrix is constructed for that order.
+%
+
+	nA = max(size(A));
+	nB = max(size(B));
+
+	n = max(nA, nB) - 1;  % order
+
+	% First create a matrix of zeros
+	E = zeros((n)*2, (n)*2);
+	
+	% Then assign specific values for A and B
+
+	% A
+	for (col = 1:n)
+		for (i = 1:nA)
+			row = (nA - i) + col;
+			E(row, col) = A(i);
+		end
+	end
+
+	% B
+	for (col = (n+1):n*2)
+		for (i = 1:nB)
+			row = (nB - i) + (col - n);
+			E(row, col) = B(i);
+		end
+	end
+endfunction

### 2013-12-05 octave/zerotime.m

@@ -1,0 +1,11 @@
+function [ts] = zerotime(tx);
+% ZEROTIME - Given a vector of incrementing times
+%  recalculate the times so that it starts at zero.
+%
+
+ts = zeros(1,length(tx));
+for (i = 2:length(tx))
+	ts(i) = (tx(i) - tx(i-1)) + ts(i-1);
+end
+
+endfunction

### 2013-12-06 octave/cd_plot2.m
finished documentation of Direct Design method
The Direct Design method has been documented which
includes the steps needed to calculate D(z) and K.

@@ -1,5 +1,5 @@
 %
-% cd2_plot.m
+% cd_plot2.m
 %
 
 clear;

### 2013-12-06 octave/dd_plot2.m

@@ -1,5 +1,5 @@
 %
-% dd_plot1.m
+% dd_plot2.m
 %
 % Plot of solution using direct design compared to no control.
 %
@@ -25,8 +25,7 @@ Gz1 = engine_model(T);
 % These roots are guess, taken from some other example.
 roots = [(0.6 + 0.4i) (0.6 - 0.4i) 0 0 0];
 order = 3;
-limit = 1;
-[Hz1, Dz1, K1] = direct_design(Gz1, roots, order, limit);
+[Hz1, Dz1, K1] = direct_design(Gz1, roots, order);
 
 % System with no control.
 Gz2 = engine_model(T);

### 2013-12-06 octave/direct_design.m

@@ -36,8 +36,7 @@ Dz = tf(Beta, Alpha, T);
 % To find K, the limit should go to 1
 % for a step input.
 % (refer to the notes for a better description)
-K = sum(D)/sum(conv(Bz, Alpha));
-% XXX
+K = limit*polyval(D, limit)/polyval(conv(Bz, Alpha), limit);
 
 Hz = K*Gz/(1 + Dz*Gz);
 

### 2013-12-06 paper/main.tex
finished documentation of Direct Design method
The Direct Design method has been documented which
includes the steps needed to calculate D(z) and K.

@@
 \tikzset{shadow/.style={drop shadow,fill=white}}
-\tikzset{sum/.style={circle,draw,very thick,shadow,minimum size=6mm}}
+\tikzset{sum/.style={circle,draw,very thick,shadow,minimum size=5mm}}
 \tikzset{sumnp/.style={sum,
@@
 
+There are several characteristics of this Direct Design method.
+The controller is built entirely in the discrete domain.
+If the plant ($G(z)$) is continuous it is converted to discrete
+using a Zero Order Hold.
+In this case it is already discrete.
+It is a pole placement technique where the desired roots of the
+entire system are specified by the designer.
+The controller values for $\beta$ and $\alpha$ are found using the
+Diophantine equation along with a Sylvester matrix.
+And the gain ($K$) is chosen such that the desired limit is reached.
+Figure \ref{fig:ddc} shows the structure of the system.
+
+% {{{ fig:ddc
+\begin{figure}[h!]
+
+\begin{center}
+\begin{tikzpicture}[>=triangle 60, scale=0.8,
+	node distance=20mm, auto]
+
+	\node[gain]		(K) 	[] 				{$K$};
+	\node[sumnp]	(sum1) 	[right of=K] 	{};
+	\node[gain]		(Gz) 	[right of=sum1]	{$G(z)$};
+	\node[gain]		(Dz) 	[below of=Gz] 	{$D(z)$};
+
+	\draw [<-] (K) -- ++(-20mm,0) node[above]{$u$};
+	\draw [->] (K) -- (sum1);
+	\draw [->] (sum1) -- (Gz);
+	\draw [<-] (sum1) |- (Dz);
+	\draw [->] (Gz) -- ++(35mm,0) node[above]{$y$};
+	\draw [->] (Gz) -- ++(20mm,0) |- (Dz.east);
+
+	% add B/A labels
+	\node [right of=Gz,node distance=0cm,xshift=9mm,yshift=6mm] {$\dfrac{B}{A}$};
+	\node [right of=Dz,node distance=0cm,xshift=9mm,yshift=6mm] {$\dfrac{\beta}{\alpha}$};
+\end{tikzpicture}
+\end{center}
+
+\caption{Direct Design system where $K$ is the scaling input, $G$ is
+the plant and $D$ is the controller.}
+\label{fig:ddc}
+\end{figure}
+% }}}
+
+In this case the roots were not given and no deterministic way
+was available to find them.
+So trial and error was used.
+
+\begin{align*}
+	\text{roots} &: z = 0.6 \pm 0.4i
+\end{align*}
+
+Since the plant is already in discrete form it is not necessary
+to use a Zero Order Hold.
+The engine model, repeated from Equation \ref{eq:lems}, is shown below.
+\begin{align*}
+G(z) &= \frac{43.38}{z^3 - 1.838 z^2 + 0.8442 z + 0.02375} \\
+	&= \frac{B(z)}{A(z)} = \frac{b_0 z^3 + b_1 z + b2}
+								{a_0 z^3 + a_1 z + a2}
+\end{align*}
+
+Next it must be arranged to place it in the form of the Diophantine Equation.
+
+From the block diagram the equation describing the system is found.
+\begin{align*}
+	\frac{y}{h} &= H \\
+	H &= K \frac{G}{1 + D G} \\
+	H &= K \frac{B \alpha}{A \alpha + B \beta}
+\end{align*}
+
+From the roots $D$ is constructed.
+\begin{align*}
+	D(z) &= (0.6 + 0.4i)(0.6 - 0.4i) \\
+	D(z) &= z^2 - 1.2z + 0.520
+\end{align*}
+
+Recall the form of the Diophantine Equation where the order is $n$.
+\[
+\begin{array}{ccccccc}
+	\alpha(z) &A(z) &+ &\beta(z) &B(z) &= &D \\
+	  (n-1)   &(n)  &  & (n-1)  &(n) & &(2n-1)
+\end{array}
+\]
+
+Since $A(z)$ is a third order polynomial $n = 3$.
+This requires that $D(z)$ be a fifth order polynomial ($2n-1$).
+The order is increased by adding zeros.
+
+\begin{align*}
+	D(z) &= z^5 - 1.2z^4 + 0.520z^3 \\
+		 &= d_0z^5 + d_1z^4 + d_2z^3 + d_3z^2 + d_4z + d_5
+\end{align*}
+
+This also requires that $\alpha$ and $\beta$ are
+second order polynomials ($n-1$).
+
+\begin{align*}
+	\frac{\beta}{\alpha} &= \frac{\beta_0 z^2 + \beta_1 z + \beta_2}
+								{\alpha_0 z^2 + \alpha_1 z + \alpha_2}
+\end{align*}
+
+To solve the Diophantine Equation
+
+\begin{align*}
+	M &= E^{-1} D \\
+	\left[
+	\begin{array}{l}
+		\alpha \\
+		\beta
+	\end{array}
+	\right] &= E^{-1} D \\
+	\left[
+	\begin{array}{l}
+		\alpha \\
+		\beta
+	\end{array}
+	\right] &= 
+	\left[
+	\begin{array}{l}
+		\alpha_2 \\
+		\alpha_1 \\
+		\alpha_0 \\
+		\beta_2 \\
+		\beta_1 \\
+		\beta_0
+	\end{array}
+	\right]
+\end{align*}
+
+a third order Sylvester Matrix is required.
+
+\[
+E=
+\begin{bmatrix}
+	a_3 & 0 & 0 & b_3 & 0 & 0 \\
+	a_2 & a_3 & 0 & b_2 & b_3 & 0 \\
+	a_1 & a_2 & a_3 & b_1 & b_2 & b_3 \\
+	a_0 & a_1 & a_2 & b_0 & b_1 & b_2 \\
+	0 & a_0 & a_1 & 0 & b_0 & b_1 \\
+	0 & 0 & a_0 & 0 & 0 & b_0 \\
+\end{bmatrix}
+\]
+
+Combining these parts and solving for $\alpha$ and $\beta$ results
+in a solution for $D(z)$.
+
+\begin{align*}
+	\Aboxed{
+	\frac{\beta}{\alpha} &= D(z) = \frac{0.02297 z^2 - 0.01686 z - 0.0004643}
+								{z^2 + 0.6379 z + 0.8482}
+	}
+\end{align*}
+
+\begin{samepage}
+The gain ($K$) still needs to be found and this can be accomplished
+using the limit.
+\begin{align*}
+\lim_{z \to 1} &K \frac{B \alpha}{A \alpha + B \beta} = 1\\
+\lim_{z \to 1} &K \frac{(43.38) (z^2 + 0.6379 z + 0.8482)}
+	{(z^5 - 1.2z^4 + 0.520z^3)} = 1 \\
+K &= \frac{(1 - 1.2 + 0.520)}{(43.38)(1 + 0.6379 + 0.8482)} \\
+\Aboxed{ K &= 0.0029671 }
+\end{align*}
+\end{samepage}
+
+This solution results in the response shown in Figure \ref{fig:dd_plot2}.
+It is a dramatic improvement compared to no control.
+The rise time, percent overshoot, and settling time are all improved.
+This result is particularly interesting because the chosen roots were a guess.
+It is likely that further improvement could be made by optimizing the roots.
+
 % {{{ fig:dd_plot2
@@
 \begin{center}
-\includegraphics[scale=0.7]{../octave/dd_plot2}
+\includegraphics[scale=0.6]{../octave/dd_plot2}
 \end{center}
@@
 compared to no control.  For no control this is the same plot
-as in Figure \ref{fig:em_plot4} which stabilizes after a second.}
+as in Figure \ref{fig:em_plot4} which stabilizes after a second.
+The Matlab source performing these calculations and producing this
+plot is given in Appendix \ref{app:dd_matlab}.}
 \label{fig:dd_plot2}
@@
 % }}}
+
+While these results were positive it is not obvious whether it is withing
+the mechanical limits of a real engine.
+In particular the control input should be limited to a range from zero to one.
+Further investigation is needed to determine whether these limits
+are being exceeded.
+There is no doubt, however, that the addition of a controller
+improved performance.
+If the limits are being exceeded a different set of roots would have to
+be chosen which may be less optimal.
 
@@
 \label{app:em_matlab}
-\label{app:rpmtime_plot_source}
 
@@
 
+% {{{ General Matlab Functions
+\clearpage
+\section{General Matlab Functions}
+\label{app:gf_matlab}
+
+This sections includes Matlab source for functions which are
+used among several methods.
+
+\lstinputlisting[
+	caption={Matlab source of function to calculate a Sylvester Matrix.} 
+	]{../octave/sylvester.m}
+
+\lstinputlisting[
+	caption={Matlab source of function to calculate zero time.} 
+	]{../octave/zerotime.m}
+
+\clearpage
+\lstinputlisting[
+	caption={Matlab source of modified lsim() function.} 
+	]{../octave/lsim1.m}
+
+% }}}
+
 % {{{ Direct Design Matlab Source
@@
 
+Matlab source code used to build a controller using Direct Design
+and plot the output.
+The \verb+engine_model+ is given in Appendix \ref{app:em_matlab}
+and other functions are given in Appendix \ref{app:gf_matlab}.
+
+\lstinputlisting[
+	caption={Matlab source of function is to build a controller using
+		Direct Design.}
+	]{../octave/direct_design.m}
+
+\clearpage
 \lstinputlisting[

### 2013-12-06 paper/main.tex
general cleanup, added contents, changed title page
The title page has been changed to only include itself
and the abstract.  On the next page is the contents and
after that is the body of the text.  Before the body
started with the title page and there was no contents.

@@
 \usepackage{sectsty}  % \sectionfont
-\sectionfont{\normalsize}
-\subsectionfont{\normalsize}
+%\sectionfont{\normalsize}
+%\subsectionfont{\normalsize}
 
@@
 
+\vspace*{1.0in}
 \centerline{\Large \textbf{Control System Design Applied to}}
@@
 \vspace{0.1in}
-\centerline{\normalsize {CSU Chico}}
-%\centerline{\today}
-%\vspace{0.1in}
+\centerline{\normalsize {California State University Chico}}
 \centerline{\small \today}
-\vspace{0.2in}
-\centerline{\LARGE \textbf{DRAFT}}
-\vspace{0.2in}
+%\vspace{0.2in}
+%\centerline{\LARGE \textbf{DRAFT}}
+\vspace{0.5in}
 
-% }}}
+\vfill
 
-% {{{ abstract
-%\pagebreak
-%\thispagestyle{empty}
 \begin{abstract}
@@
 various state space designs.
+\end{abstract}
 
-\end{abstract}
+\vspace{1.5in}
+
 % }}}
 
+\clearpage
+\tableofcontents
+
 % {{{ Engine Model
+\clearpage
 \section{Engine Model}
@@
 
-While these results were positive it is not obvious whether it is withing
+While this result is positive it is not apparent whether it is within
 the mechanical limits of a real engine.
@@
 
-The following code has been tested using Octave\autocite{octave},
-an open source Matlab clone.
+\nocite{octave}
 
@@
 
-For a given control input with zero torque the stable rpm values reached
-are shown in Figure \ref{fig:stable}.
+For a given control input and zero torque the engine model will
+reach stable rpm values as shown in Figure \ref{fig:stable}.
 Equation \ref{eq:stable} describes its behavior.

### 2013-12-13 octave/cd_plot1.m
changed print -color -deps2 to -depsc2

@@ -28,4 +28,4 @@ lsim(sys, u);
 grid on;
 axis auto;
 
-print('cd_plot1.eps', '-color', '-deps2');
+print('cd_plot1.eps', '-depsc2');

### 2013-12-13 octave/cd_plot2.m

@@ -24,4 +24,4 @@ lsim(sys, u);
 grid on;
 axis auto;
 
-print('cd_plot2.eps', '-color', '-deps2');
+print('cd_plot2.eps', '-depsc2');

### 2013-12-13 octave/dc_plot1.m

@@ -28,4 +28,4 @@ lsim(sys, u);
 grid on;
 axis auto;
 
-print('dc_plot1.eps', '-color', '-deps2');
+print('dc_plot1.eps', '-depsc2');

### 2013-12-13 octave/dd_plot1.m

@@ -46,4 +46,4 @@ title(sprintf('Output Response, step: %d - %d rpm', u_lo, u_hi));
 ylabel('rpm');
 xlabel('time (sec)');
 
-print('dd_plot1.eps', '-color', '-deps2');
+print('dd_plot1.eps', '-depsc2');

### 2013-12-13 octave/dd_plot2.m

@@ -68,4 +68,4 @@ ylabel('rpm');
 xlabel('time (sec)');
 legend('no control', 'direct design');
 
-print('dd_plot2.eps', '-color', '-deps2');
+print('dd_plot2.eps', '-depsc2');

### 2013-12-13 octave/em_plot1.m

@@ -30,4 +30,4 @@ title('Output Response');
 ylabel('rpm');
 xlabel('ticks');
 
-print('em_plot1.eps', '-color', '-deps2');
+print('em_plot1.eps', '-depsc2');

### 2013-12-13 octave/em_plot3.m

@@ -31,4 +31,4 @@ title('Stable Output Rpm');
 xlabel('control input');
 ylabel('rpm');
 
-print('em_plot3.eps', '-color', '-deps2');
+print('em_plot3.eps', '-depsc2');

### 2013-12-13 octave/em_plot4.m

@@ -36,4 +36,4 @@ title('Engine Response With Control From 0.41 to 0.69');
 xlabel('time (sec)');
 ylabel('rpm');
 
-print('em_plot4.eps', '-color', '-deps2');
+print('em_plot4.eps', '-depsc2');

### 2013-12-13 octave/dc_plot2.m
changed print -color -deps2 to -depsc2

@@ -26,4 +26,4 @@ grid on;
 axis auto;
 axis([0 7 -4 6]);
 
-print('dc_plot2.eps', '-color', '-deps2');
+print('dc_plot2.eps', '-depsc2');

### 2013-12-13 octave/em_plot2.m

@@ -43,4 +43,4 @@ title('Output Response');
 ylabel('rpm');
 xlabel('time (sec)');
 
-print('em_plot2.eps', '-color', '-deps2');
+print('em_plot2.eps', '-depsc2');

### 2013-12-16 paper/main.tex
regulator system is wrong size?
An attempt was made to construct a regulator system
(Control Law, Full Order) however the sizes do not match.
The system (Gz) has one input and one output however the
feedback (K), as produced using Ackermann's Formula
has three inputs for the 3rd order system.
Is this a case where the other values must be estimated?

@@
 \clearpage
-\section{Regulator Control System}
-
-What is needed is a regulator system where the input is applied
-as torque.
-The general system is shown in Figure \ref{fig:rpm1}.
-
-% {{{ fig:rpm1, Regulator System
+\section{Regulator Control System, Control Law}
+
+Instead of a servo system, which accepts a control input,
+a regulator system can be constructed which tries to maintain
+a single output value.  The general design is shown
+in Figure \ref{fig:reg1}.
+
+% {{{ fig:reg1, Regulator System
 \begin{figure}[hbp!]
@@
 \end{center}
-\caption{General system for regulating engine rpm.}\label{fig:rpm1}
+\caption{Regulator system for idle rpm control.}\label{fig:reg1}
+\end{figure}
+% }}}
+
+One way to build this regulator system is to use the
+Control Law Design\autocite[Pg. 280]{franklin1998digital}.
+The block diagram of the system is shown in Figure \ref{fig:clff1}.
+
+% {{{ fig:clff1
+\begin{figure}[hpb!]
+\begin{center}
+
+\begin{tikzpicture}[>=triangle 60,
+	node distance=14mm, auto]
+
+	\node[gain]  	(Gz)	[] 	{$\begin{matrix}
+									\dot{x} = Ax + Bu \\
+										 y = Cx + Du
+								 \end{matrix}$};
+	\node[gain] 	(K)	[below=of Gz] 	{$-K$};
+
+	 \draw [->]
+	 		(Gz.east) -- ++(15mm,0)
+			-| ++(0,-10mm)
+			|- (K.east);
+
+	 \draw [<-]
+	 		(Gz.west) -- ++(-15mm,0)
+			-| ++(0,-10mm)
+			|- (K.west);
+
+\end{tikzpicture}
+
+\end{center}
+
+\caption{Control Law system with full feedback.
+It behaves as a regulator since not input is available.}
+\label{fig:clff1}
 \end{figure}

### 2013-12-16 octave/cl_ff01.m

@@ -1,0 +1,68 @@
+%
+% cl_ff01.m
+%
+% Control Law with full feedback.
+%
+
+clear;
+clf;
+
+T = 1;  % time step
+Tend = 10;
+
+% roots
+z1 = [0.9 0 0];
+%z1 = [0.9 0.9 0];
+%z2 = [(0.9 + 0.05i) (0.9 - 0.5i) 0];
+%z1 = [0.9 0.9 0.9 0.9];
+%z2 = [(0.9 + 0.05i) (0.9 - 0.5i) (0.8 + 0.4i) (0.8 - 0.4i)];
+
+Gz = engine_model(T);
+
+[A,B,C,D] = tf2ss(Gz);
+sysD = ss(Gz);
+
+n = length(A);  % order
+
+[Phi,Gamma,H,G] = ssdata(sysD);
+
+% can't use 'place' with identical roots
+%K1 = place(Phi, Gamma, z1);
+K1 = myacker(Phi, Gamma, z1);
+%K2 = myacker(Phi, Gamma, z2);
+
+% create a gain of K using D and zeroing others
+sysK1 = ss(zeros(1,1), zeros(1,3), zeros(1,1), K1, T);
+%sysK2 = ss(zeros(1,1), zeros(1,n), zeros(1,1), K2, T);
+
+sysCL1 = feedback(sysD, sysK1, -1);
+%sysCL2 = feedback(sysD, sysK2, -1);
+
+% simplify, structural pole/zero cancellation
+sysCL1 = sminreal(sysCL1);
+%sysCL2 = sminreal(sysCL2);
+
+%u = zeros([(Tend/T) 1]);
+%[y1,t1] = lsim(sysCL1, u, [], x0);
+%[y2,t2] = lsim(sysCL2, u, [], x0);
+
+%figure(1);
+%subplot(2, 1, 1);
+%%plot(t2, y1);  % all
+%plot(t2, y1(:,1));  % one
+%grid on;
+%axis tight;
+%title('Regulator Response, roots: (0.9 0.9 0.9 0.9)');
+%%xlabel('time (sec)');
+%ylabel('y');
+%
+%subplot(2, 1, 2);
+%%plot(t2, y2);  % all
+%plot(t2, y2(:,1));  % one
+%grid on;
+%axis tight;
+%title('Regulator Response, roots: (0.9 + 0.05i) (0.9 - 0.5i) (0.8 + 0.4i) (0.8 - 0.4i)');
[…]

### 2013-12-16 octave/myacker.m

@@ -1,0 +1,49 @@
+
+function [K] = myacker(A, B, z)
+% MYACKER An implementation of ACKER.
+%
+% MYACKER is an implementation of the Matlab ACKER function.
+%
+% Given a feedback system it is necessary to find K such
+% that the regulator will go to zero.
+%
+%        +--------------+
+%    +-->| x' = Ax + Bu |---+
+%    |   +--------------+   |
+%    |                      |
+%    |       +----+         |
+%    +-------| -K |<--------+
+%            +----+
+%
+% By specifying the desired roots (z) K can be found
+% using Ackermann's Formula.
+%
+% The number of roots must match the order of the system.
+% And the choice of roots determines how well the system performs.
+%
+% See also ACKER, PLACE.
+
+n = size(A, 1);  % order of problem
+zp = poly(z);
+
+% Controllability Matrix
+% [B AB A^2B ... A^(n-1)B]
+W = ctrb(A, B);
+
+% [0 0 ... 1]
+en = zeros([1 n]);
+en(n) = 1;
+
+% alpha = A^n + z(1)*A^(n-1) + ... + z(n)*I
+alpha = 0;
+for i = 0:n
+    if (i == n)
+        % last
+        alpha = alpha + zp(i+1)*eye(n);
+    else
+        alpha = alpha + zp(i+1)*A^(n-i);
+    end
+end
+
+% z = eig(A - B*K)
+K = real(en*(W\alpha));

### 2013-12-17 octave/cl_ff01.m
all args to lsim1, regulator still isn't right
lsim1 has been expanded to support all the arguments.

The regulator system still doesn't work correctly and it
hasn't been determined why.  It needs an offset since
the regulator goes towards zero but the rpm should stabilize
at 700 rpm.  And the zeros chosen for K have a large impact
on the result.  These should influence how good the result
is, not whether it works at all.

@@ -8,61 +8,65 @@ clear;
 clf;
 
 T = 1;  % time step
-Tend = 10;
-
-% roots
-z1 = [0.9 0 0];
-%z1 = [0.9 0.9 0];
-%z2 = [(0.9 + 0.05i) (0.9 - 0.5i) 0];
-%z1 = [0.9 0.9 0.9 0.9];
-%z2 = [(0.9 + 0.05i) (0.9 - 0.5i) (0.8 + 0.4i) (0.8 - 0.4i)];
+Tend = 100;
+ncyl = 8;  % number of cylinders
+rpm = 700;  % idle rpm
 
 Gz = engine_model(T);
+[Phi,Gamma,H,G] = ssdata(Gz);
 
-[A,B,C,D] = tf2ss(Gz);
-sysD = ss(Gz);
+n = length(Phi);  % order
 
-n = length(A);  % order
-
-[Phi,Gamma,H,G] = ssdata(sysD);
-
-% can't use 'place' with identical roots
-%K1 = place(Phi, Gamma, z1);
+% Find K
+% roots
+z1 = [0.9 0.9 0];
+%z1 = [0.5 0.5 0];
+%z1 = [(0.9 + 0.05i) (0.9 - 0.5i) 0];
+%z1 = [(0.5 + 0.05i) (0.5 - 0.5i) 0];
+%z1 = [0.9 0.9 0.9 0.9];
 K1 = myacker(Phi, Gamma, z1);
-%K2 = myacker(Phi, Gamma, z2);
-
 % create a gain of K using D and zeroing others
-sysK1 = ss(zeros(1,1), zeros(1,3), zeros(1,1), K1, T);
-%sysK2 = ss(zeros(1,1), zeros(1,n), zeros(1,1), K2, T);
+K1z = ss(zeros(1,1), zeros(1,n), zeros(1,1), K1, T);
 
-sysCL1 = feedback(sysD, sysK1, -1);
-%sysCL2 = feedback(sysD, sysK2, -1);
+% Find Predictor Estimator (Lp)
+%H = [1 0 0];  % filter
+% roots
+z2 = z1*0.84;
+%z2 = [(0.4 + 0.4i) (0.4 - 0.4i) 0];
+%Lp = place(Phi', H', z2)';
+Lp = myacker(Phi', H', z2)';
+Pz = ss((Phi - Gamma*K1 - Lp*H), Lp, eye(n), zeros(n,1), T);
 
-% simplify, structural pole/zero cancellation
[…]

### 2013-12-17 octave/lsim1.m
all args to lsim1, regulator still isn't right
lsim1 has been expanded to support all the arguments.

The regulator system still doesn't work correctly and it
hasn't been determined why.  It needs an offset since
the regulator goes towards zero but the rpm should stabilize
at 700 rpm.  And the zeros chosen for K have a large impact
on the result.  These should influence how good the result
is, not whether it works at all.

@@ -1,10 +1,10 @@
-function [y, t, x, i] = lsim1(Hz, u);
+function [y, t, x, i] = lsim1(Hz, u, t=[], x0=false);
 % LSIM1 - Just like lsim excpt small values are removed.
 %
 % "small" means less than or equal to 10.
 %
 
-[y, t, x] = lsim(Hz, u);
+[y, t, x] = lsim(Hz, u, t, x0);
 
 % Find the first valid value
 for i = 1:length(y) 

### 2013-12-17 octave/lsim1.m
removed endfunction for compatibility with Matlab

@@ -20,5 +20,3 @@ u = u(i:end);
 t = t(i:end);
 
 i = i - 1;
-
-endfunction

### 2013-12-17 octave/sylvester.m

@@ -30,4 +30,3 @@ function [E] = sylvester(A, B)
 			E(row, col) = B(i);
 		end
 	end
-endfunction

### 2013-12-17 octave/zerotime.m

@@ -7,5 +7,3 @@ ts = zeros(1,length(tx));
 for (i = 2:length(tx))
 	ts(i) = (tx(i) - tx(i-1)) + ts(i-1);
 end
-
-endfunction

### 2013-12-17 octave/engine_model.m
removed endfunction for compatibility with Matlab

@@ -15,5 +15,3 @@ D = tf([0.00093], [1], T);
 H = K*G/(1 + D*G);
 
 sys = minreal(H);  % cancel common roots
-
-endfunction

### 2013-12-17 matlab/linear_engine_model_plot.m
removed matlab specific files (to own branch)

@@ -1,34 +1,0 @@
-clear all;
-clf;
-
-linear_engine_model_init;
-
-sim('linear_engine_model');
-
-figure;
-
-
-
-subplot(3,1,1);
-plot(dout.Time, dout.Data(:,1));
-grid on;
-axis tight;
-title('Idle Duty');
-ylabel('idle duty (%)');
-xlabel('time (sec)');
-
-subplot(3,1,2);
-plot(dout.Time, dout.Data(:,3));
-grid on;
-axis tight;
-title('Torque');
-ylabel('torque');
-xlabel('time (sec)');
-
-subplot(3,1,3);
-plot(dout.Time, dout.Data(:,2));
-grid on;
-axis tight;
-title('RPM');
-ylabel('RPM');
-xlabel('time (sec)');

### 2013-12-17 matlab/linear_engine_model_init.m
removed matlab specific files (to own branch)

@@ -1,16 +1,0 @@
-% initialization for the Linear Engine Model
-% 'linear_engine_model.mdl'
-
-idle_rpm = 700;  % (RPM)
-num_cyl = 8;
-% frequency of ignition impulses
-% In two crankshaft rotations all cylinders will fire
-% for a 4-stroke engine.
-idle_rps = 700/60;  % minute -> second
-ign_freq = num_cyl*idle_rps*0.5;  % (Hz)
-
-% engine model time step
-Tm = 1/ign_freq;  % (sec)
-
-% control time step
-Tc = 0.5;  % (sec)

### 2013-12-17 octave/cl_ff01.m
adding an offset to the tf doesn't even work?

@@ -13,15 +13,19 @@ ncyl = 8;  % number of cylinders
 rpm = 700;  % idle rpm
 
 Gz = engine_model(T);
-[Phi,Gamma,H,G] = ssdata(Gz);
+Gzx = Gz - rpm;
+% XXX - Gzx should output zero at rpm, but it doesn't
+% A control input of 47/1450.1 results in 700 rpm. Why?
+[Phi,Gamma,H,G] = ssdata(Gzx);
+% XXX - is this the right H?
 
 n = length(Phi);  % order
 
 % Find K
 % roots
-z1 = [0.9 0.9 0];
+%z1 = [0.9 0.9 0];
 %z1 = [0.5 0.5 0];
-%z1 = [(0.9 + 0.05i) (0.9 - 0.5i) 0];
+z1 = [(0.9 + 0.05i) (0.9 - 0.5i) 0];
 %z1 = [(0.5 + 0.05i) (0.5 - 0.5i) 0];
 %z1 = [0.9 0.9 0.9 0.9];
 K1 = myacker(Phi, Gamma, z1);
@@ -52,7 +56,7 @@ sysX = sminreal(sysX);
 % Simulate
 x0 = [1; zeros(5,1)];
 u1 = 0*ones([Tend/T 1]);
-u2 = -rpm*ones([Tend/T 1]);
+u2 = -rpm*ones([Tend/T 1]);  % XXX - shouldn't be negative!
 u = [u1 u2];
 [y1,t1] = lsim1(sysX, u, [], x0);
 %y1 = y1 + rpm;

### 2013-12-18 octave/cl_ff01.m
got a Control Law example working, by intuition...
Difficulties have been encountered getting a Control Law
example working due to its particular requirements and
the characteristics of the engine model.  Specifically
the control input should reflect the output and it should
stabilize at zero for negative feedback to work.

An example has been designed intuitively.  Simply by making
the correct compensation to G(z) and choosing some K to
be feed back.  It is not the fastest, but it works and it
is stable.

@@ -2,75 +2,64 @@
 % cl_ff01.m
 %
 % Control Law with full feedback.
+% Intuitive Design.
 %
 
 clear;
-clf;
 
 T = 1;  % time step
-Tend = 100;
+Tend = 200;  % end in ticks
 ncyl = 8;  % number of cylinders
-rpm = 700;  % idle rpm
+% stabilizes at rpm_lo then steps to rpm_hi
+rpm_hi = 700;  % idle rpm
+rpm_lo = 200;
+
+% choose any K <= 1
+% larger values increase overshoot
+K = 0.01;
 
 Gz = engine_model(T);
-Gzx = Gz - rpm;
-% XXX - Gzx should output zero at rpm, but it doesn't
-% A control input of 47/1450.1 results in 700 rpm. Why?
-[Phi,Gamma,H,G] = ssdata(Gzx);
-% XXX - is this the right H?
-
-n = length(Phi);  % order
-
-% Find K
-% roots
-%z1 = [0.9 0.9 0];
-%z1 = [0.5 0.5 0];
-z1 = [(0.9 + 0.05i) (0.9 - 0.5i) 0];
-%z1 = [(0.5 + 0.05i) (0.5 - 0.5i) 0];
-%z1 = [0.9 0.9 0.9 0.9];
-K1 = myacker(Phi, Gamma, z1);
-% create a gain of K using D and zeroing others
-K1z = ss(zeros(1,1), zeros(1,n), zeros(1,1), K1, T);
-
-% Find Predictor Estimator (Lp)
-%H = [1 0 0];  % filter
-% roots
-z2 = z1*0.84;
-%z2 = [(0.4 + 0.4i) (0.4 - 0.4i) 0];
-%Lp = place(Phi', H', z2)';
-Lp = myacker(Phi', H', z2)';
-Pz = ss((Phi - Gamma*K1 - Lp*H), Lp, eye(n), zeros(n,1), T);
 
 % Build System
 % upper branch
-set(Gz, 'inname', 'gu', 'outname', 'gy');
-sum1 = sumblk('y = gy - rpm');
-Gzs = connect(Gz, sum1, {'gu', 'rpm'}, {'y'});
-% lower feedback branch
-K1Pz = series(Pz, K1);
-set(K1Pz, 'inname', 'u1', 'outname', 'y1');
+sum1 = sumblk('e1 = u + r1');
+% convert rpm to control
[…]

### 2013-12-18 paper/main.tex
moved Control Response section to start of Appendix
documented Control Law intuitive design
The Control Law intuitive design has been fully documented
with block diagrams and plots.

--- 2013-12-06 paper/main.tex
@@
 				label={{above right,xshift=-0.0mm,yshift=-1.0mm}:$+$}} }
+\tikzset{sumxpxn/.style={sum,
+				label={{above left,xshift=0.0mm,yshift=-1.0mm}:$+$},
+				label={{above right,xshift=-0.0mm,yshift=-1.0mm}:$-$}} }
 \tikzset{gain/.style={rectangle,draw,very thick,inner sep=2mm,shadow}}
@@
 
-% {{{ Regulator Control System
-\clearpage
-\section{Regulator Control System}
-
-What is needed is a regulator system where the input is applied
-as torque.
-The general system is shown in Figure \ref{fig:rpm1}.
-
-% {{{ fig:rpm1, Regulator System
-\begin{figure}[hbp!]
-\begin{center}
-
-\begin{tikzpicture}[>=triangle 60, scale=0.8,
+% {{{ Control Law, Intuitive Design
+\clearpage
+\section{Control Law, Intuitive Design}
+
+% {{{ fig:clid01
+\begin{figure}[hpb!]
+\begin{center}
+
+\begin{tikzpicture}[>=triangle 60,
 	node distance=14mm, auto]
 
-	\node[gain] 	(engine)	[]					{engine};
-	\node[gain] 	(control)	[below=of engine]	{controller};
-
-	\draw [<-] (engine.north) -- ++(0,10mm) node[above]{$\text{torque}$};
-	\draw [->] (engine.east) node[above,xshift=7mm]{rpm}
-					-- ++(20mm,0)
-					|- (control.east);
-	\draw [->] (control.west)
-					-- ++(-20mm,0)
-					|- (engine.west) node[above,xshift=-9mm]{control};
+	\node[gain]  	(Gz)	[] 	{$G(z)$};
+	\node[gain] 	(K)	[below=of Gz] 	{$K$};
+
+	 \draw [->]
+	 		(Gz.east) -- ++(15mm,0)
+			-| ++(0,-10mm)
+			|- (K.east);
+
+	 \draw [<-]
+	 		(Gz.west) -- ++(-15mm,0)
+			-| ++(0,-10mm)
+			|- (K.west);
 \end{tikzpicture}
@@
 \end{center}
-\caption{General system for regulating engine rpm.}\label{fig:rpm1}
-\end{figure}
+
+\caption{Control Law regulator system.
+In typical designs a negative feedback is produced using a negative $K$.}
+\label{fig:clid01}
+\end{figure}
+% }}}
+
+A Control Law Design\autocite[Pg. 280]{franklin1998digital} configured
+as a regulator is shown in Figure \ref{fig:clid01}.
+Notice that there is no input causing it regulate at a constant set point.
+However there are several requirements that must be met in order
+for this system to be stable.
+First the input to $G(z)$ must reflect the output without any offset.
+And second the system will always stabilize at zero.
+
+These requirements can be verified intuitively.
+Suppose for an input of $1$ to $G(z)$ and output of $5$ is produced.
+What value of $K$ will produce a stable value?
+
+\begin{align*}
+	5 K &= 1 \\
+	K &= \frac{1}{5}
+\end{align*}
+
+Now suppose a disturbance is introduced causing the output to increase
+to $6$, will the system stabilize?
+
+\begin{align*}
+	6 \cdot K &= y \\
+	6 \cdot \frac{1}{5} &= \frac{6}{5} \\
+	\frac{6}{5} &> 1 \quad \text{(diverge)}
+\end{align*}
+
+An increase in the output creates positive feedback which increases
+the input and causes the system diverge.
+This exercise can be repeated with a controller that has no offset at zero.
+The result is a negative feedback system that will stabilize at zero.
+
+The linear engine model with zero torque (Equation \ref{eq:lems})
+supports none of these requirements for a stable Control Law regulator.
+The input does not reflect the output; the input range is from zero to
+one whereas the output range is from 200 to 1200.
+And operation at zero rpm is not realistic.
+
+% {{{ fig:clid02
+\begin{figure}[hpb!]
+\begin{center}
+
+\begin{tikzpicture}[>=triangle 60,
+	node distance=14mm, auto]
+
+	\node[sumxpxp]  	(sum1)	[]				{};
+	\node[gain]  	(R)		[right=of sum1] 	{$R$};
+	\node[gain]  	(Gz)	[right=of R] 		{$G(z)$};
+	\node[sumxpxn]	(sum2)	[right=of Gz]		{};
+	\node[gain] 	(K)		[below=of Gz] 		{$-K$};
+
+	\draw [<-] (sum1.north) -- ++(0,10mm) node[above]{$\text{rpm}$};
+	\draw [->] (sum1.east) -- (R.west);
+	\draw [->] (R.east) -- (Gz.west);
+	\draw [->] (Gz.east) -- (sum2.west);
+	\draw [<-] (sum2.north) -- ++(0,10mm) node[above]{$\text{rpm}$};
+	\draw [->] (sum2.east) -- ++(10mm,0) node[above,xshift=-3mm]{$y$}
+		-| ++(0,-10mm)
+		|- (K.east);
+	 \draw [<-] 
+	 		(sum1.west) -- ++(-10mm,0) node[above,xshift=3mm]{$u$}
+			|- (K.west);
+\end{tikzpicture}
+
+\end{center}
+
+\caption{Compensations added to the linear engine model with zero
+torque ($G(z)$) to construct a regulator system.
+The transfer function $R$ converts rpm input to control output (0 - 1).
+The rpm input is the desired idle speed such as 700.
+Notice that both the output ($y$) and input ($u$) will be zero when
+the desired rpm is reached.}
+\label{fig:clid02}
+\end{figure}
+% }}}
+
+The system in Figure \ref{fig:clid02} adds compensations to the
+engine model so that a regulator system can be constructed.
+Figure \ref{fig:cl_plot1} shows the response of this system.
+While the response isn't very good the design was trivial
+by simply choosing an arbitrary constant of $0.01$ for $K$.
+Later designs will improve performance by elaborating upon this
+system and developing more complex values for $K$.
+
+% {{{ fig:dd_plot2
+\begin{figure}[htbp!]
+\begin{center}
+\includegraphics[scale=0.6]{../octave/cl_ff01}
+\end{center}
+\caption{Output response of Control Law regulator with $K = 0.01$
+and rpm = $700$.
+Rpm was allowed to stabilize at 200 rpm before applying a step
+to 700 rpm.
+Source code is provided in Appendix \ref{app:cl_matlab}.}
+\label{fig:cl_plot1}
+\end{figure}
+% }}}
+
+%Instead of a servo system, which accepts a control input,
+%a regulator system can be constructed which tries to maintain
+%a single output value.  The general design is shown
+%in Figure \ref{fig:reg1}.
+
+% {{{ fig:reg1, Regulator System
+%\begin{figure}[hbp!]
+%\begin{center}
+%
+%\begin{tikzpicture}[>=triangle 60, scale=0.8,
+%	node distance=14mm, auto]
+%
+%	\node[gain] 	(engine)	[]					{engine};
+%	\node[gain] 	(control)	[below=of engine]	{controller};
+%
+%	\draw [<-] (engine.north) -- ++(0,10mm) node[above]{$\text{torque}$};
+%	\draw [->] (engine.east) node[above,xshift=7mm]{rpm}
+%					-- ++(20mm,0)
+%					|- (control.east);
+%	\draw [->] (control.west)
+%					-- ++(-20mm,0)
+%					|- (engine.west) node[above,xshift=-9mm]{control};
+%\end{tikzpicture}
+%
+%\end{center}
+%\caption{Regulator system for idle rpm control.}\label{fig:reg1}
+%\end{figure}
+% }}}
+
+%One way to build this regulator system is to use the
+%Control Law Design\autocite[Pg. 280]{franklin1998digital}.
+%The block diagram of the system is shown in Figure \ref{fig:clff1}.
+
+% {{{ fig:clff1
+%\begin{figure}[hpb!]
+%\begin{center}
+%
+%\begin{tikzpicture}[>=triangle 60,
+%	node distance=14mm, auto]
+%
+%	\node[gain]  	(Gz)	[] 	{$\begin{matrix}
+%									\dot{x} = Ax + Bu \\
+%										 y = Cx + Du
+%								 \end{matrix}$};
+%	\node[gain] 	(K)	[below=of Gz] 	{$-K$};
+%
+%	 \draw [->]
+%	 		(Gz.east) -- ++(15mm,0)
+%			-| ++(0,-10mm)
+%			|- (K.east);
+%
+%	 \draw [<-]
+%	 		(Gz.west) -- ++(-15mm,0)
+%			-| ++(0,-10mm)
+%			|- (K.west);
+%
+%\end{tikzpicture}
+%
+%\end{center}
+%
+%\caption{Control Law system with full feedback.
+%It behaves as a regulator since not input is available.}
+%\label{fig:clff1}
+%\end{figure}
 % }}}
@@
 
+% {{{ Control Response
+\clearpage
+\section{Control Response}
+\label{app:contrpm}
+
+For a given control input and zero torque the engine model will
+reach stable rpm values as shown in Figure \ref{fig:stable}.
+Equation \ref{eq:stable} describes its behavior.
+Listing \ref{lst:stable} shows how this was calculated.
+
+\begin{align}
+y &= 1450.1 x \label{eq:stable} \\
+	y &: \text{rpm (200 - 1200)} \notag \\
+	x &: \text{control (0 - 1)} \notag
+\end{align}
+
+\begin{figure}[htbp!]
+\begin{center}
+\includegraphics[scale=0.5]{../octave/em_plot3}
+\end{center}
+\caption{Stable rpm reached for a given constant control input.}
+\label{fig:stable}
+\end{figure}
+
+\clearpage
+\lstinputlisting[
+	caption={Matlab source to find the stable rpm for a given
+		control input.},
+	label={lst:stable}
+	]{../octave/em_plot3.m}
+% }}}
+
 % {{{ Engine Model Matlab Source
@@
 
-% {{{ Control Response
-\clearpage
-\section{Control Response}
-\label{app:contrpm}
-
-For a given control input and zero torque the engine model will
-reach stable rpm values as shown in Figure \ref{fig:stable}.
-Equation \ref{eq:stable} describes its behavior.
-Listing \ref{lst:stable} shows how this was calculated.
-
-\begin{align}
-y &= 1450.1 x \label{eq:stable}
-\end{align}
-
-\begin{figure}[htbp!]
-\begin{center}
-\includegraphics[scale=0.5]{../octave/em_plot3}
-\end{center}
-\caption{Stable rpm reached for a given constant control input.}
-\label{fig:stable}
-\end{figure}
-
-\clearpage
+% {{{ Control Law Matlab Source
+\clearpage
+\section{Control Law Matlab Source}
+\label{app:cl_matlab}
+
+Matlab source code used to the Control Law regulator.
+
 \lstinputlisting[
-	caption={Matlab source to find the stable rpm for a given
-		control input.},
-	label={lst:stable}
-	]{../octave/em_plot3.m}
+	caption={Matlab source used to plot the Control Law regulator.}
+	]{../octave/cl_ff01.m}
+
 % }}}

### 2013-12-18 octave/cl_ff01.m
documented Control Law intuitive design
The Control Law intuitive design has been fully documented
with block diagrams and plots.

@@ -1,8 +1,7 @@
 %
 % cl_ff01.m
 %
-% Control Law with full feedback.
-% Intuitive Design.
+% Control Law, Intuitive Design.
 %
 
 clear;

### 2013-12-19 paper/main.tex
cleaned up Control Law, Intuitive Design section
General cleanup of the wording for the Control Law,
Intuitive Design section.  And removed some old
disabled sections.

@@
 \caption{Control Law regulator system.
-In typical designs a negative feedback is produced using a negative $K$.}
+Typically, $-K$ is used to create negative feedback.}
 \label{fig:clid01}
@@
 as a regulator is shown in Figure \ref{fig:clid01}.
-Notice that there is no input causing it regulate at a constant set point.
-However there are several requirements that must be met in order
-for this system to be stable.
-First the input to $G(z)$ must reflect the output without any offset.
-And second the system will always stabilize at zero.
-
+Notice that there is no input which means it acts as a regulator as
+opposed to a servo.
+
+There are several requirements that must be met in order for this
+system to be stable.
+First, the input to $G(z)$ must reflect the output without any offset.
+And second the system must stabilize at zero.
 These requirements can be verified intuitively.
-Suppose for an input of $1$ to $G(z)$ and output of $5$ is produced.
+Suppose for an input of $1$ to $G(z)$ an output of $5$ is produced.
 What value of $K$ will produce a stable value?
@@
 An increase in the output creates positive feedback which increases
-the input and causes the system diverge.
+the input causing a further increase in the output.
+The system diverges.
 This exercise can be repeated with a controller that has no offset at zero.
@@
 The linear engine model with zero torque (Equation \ref{eq:lems})
-supports none of these requirements for a stable Control Law regulator.
+satisfies neither of these requirements.
 The input does not reflect the output; the input range is from zero to
@@
 % {{{ fig:clid02
-\begin{figure}[hpb!]
+\begin{figure}[!htp]
 \begin{center}
@@
 The transfer function $R$ converts rpm input to control output (0 - 1).
-The rpm input is the desired idle speed such as 700.
+The rpm input is the desired idle speed (e.g. 700).
 Notice that both the output ($y$) and input ($u$) will be zero when
@@
 engine model so that a regulator system can be constructed.
-Figure \ref{fig:cl_plot1} shows the response of this system.
-While the response isn't very good the design was trivial
-by simply choosing an arbitrary constant of $0.01$ for $K$.
-Later designs will improve performance by elaborating upon this
-system and developing more complex values for $K$.
-
-% {{{ fig:dd_plot2
-\begin{figure}[htbp!]
+An input of zero should produce the desired rpm.
+And this rpm input is converted to a control value ($R$).
+And the reached rpm (output of $G(z)$) is subtracted from
+the goal rpm to produce a zero output.
+
+% {{{ fig:cl_plot1
+\begin{figure}[!htp]
 \begin{center}
@@
 
-%Instead of a servo system, which accepts a control input,
-%a regulator system can be constructed which tries to maintain
-%a single output value.  The general design is shown
-%in Figure \ref{fig:reg1}.
-
-% {{{ fig:reg1, Regulator System
-%\begin{figure}[hbp!]
-%\begin{center}
-%
-%\begin{tikzpicture}[>=triangle 60, scale=0.8,
-%	node distance=14mm, auto]
-%
-%	\node[gain] 	(engine)	[]					{engine};
-%	\node[gain] 	(control)	[below=of engine]	{controller};
-%
-%	\draw [<-] (engine.north) -- ++(0,10mm) node[above]{$\text{torque}$};
-%	\draw [->] (engine.east) node[above,xshift=7mm]{rpm}
-%					-- ++(20mm,0)
-%					|- (control.east);
-%	\draw [->] (control.west)
-%					-- ++(-20mm,0)
-%					|- (engine.west) node[above,xshift=-9mm]{control};
-%\end{tikzpicture}
-%
-%\end{center}
-%\caption{Regulator system for idle rpm control.}\label{fig:reg1}
-%\end{figure}
-% }}}
-
-%One way to build this regulator system is to use the
-%Control Law Design\autocite[Pg. 280]{franklin1998digital}.
-%The block diagram of the system is shown in Figure \ref{fig:clff1}.
-
-% {{{ fig:clff1
-%\begin{figure}[hpb!]
-%\begin{center}
-%
-%\begin{tikzpicture}[>=triangle 60,
-%	node distance=14mm, auto]
-%
-%	\node[gain]  	(Gz)	[] 	{$\begin{matrix}
-%									\dot{x} = Ax + Bu \\
-%										 y = Cx + Du
-%								 \end{matrix}$};
-%	\node[gain] 	(K)	[below=of Gz] 	{$-K$};
-%
-%	 \draw [->]
-%	 		(Gz.east) -- ++(15mm,0)
-%			-| ++(0,-10mm)
-%			|- (K.east);
-%
-%	 \draw [<-]
-%	 		(Gz.west) -- ++(-15mm,0)
-%			-| ++(0,-10mm)
-%			|- (K.west);
-%
-%\end{tikzpicture}
-%
-%\end{center}
-%
-%\caption{Control Law system with full feedback.
-%It behaves as a regulator since not input is available.}
-%\label{fig:clff1}
-%\end{figure}
-% }}}
+Figure \ref{fig:cl_plot1} shows the response of this system.
+While the response isn't very good the design was trivial.
+A value for $K$ was arbitrarily chosen.
+Any values less than one were stable and larger values diverged.
+Later designs will improve performance by elaborating upon this
+system and developing more complex values for $K$.
 

### 2013-12-19 paper/main.tex
added Colophon
A Colophon has been added that describes how this
document was designed.  The References section has
also moved to the end of the document.

@@
 
-% References
-\clearpage
-\printbibliography[heading=bibintoc]
+% TODO
 
@@
 
+% {{{ Colophon
+\clearpage
+\section{Colophon}
+\label{app:colophon}
+
+This written in \LaTeX and built using Pdflatex.
+The block diagrams were written using TikZ\autocite{tikz}.
+All calculations and plots were run using version 3.6.4 of
+Octave\autocite{octave}, an open source Matlab clone.
+When possible the examples were tested in Matlab/Simulink.
+However much of the code is dependent upon the
+Octave Control (CACSD) Tools\autocite{octave-control} which are
+not entirely compatible with Matlab.
+This document and all of its code are released under the
+GNU General Public License and freely
+available\autocite{github_jmahler_idle-control}.
+\nocite{github_jmahler_control}
+
+% }}}
+
+% References
+\clearpage
+\printbibliography[heading=bibintoc]
+
 \end{document}

### 2013-12-19 octave/cl_fonp02.m
Control Law on engine model still isn't working
Several attempts have been made to try and get
a Control Law with full feedback working with the
engine model.  But the non-zero stable point and
input/output offsets have created difficulties
that haven't be resolved yet.

@@ -1,0 +1,78 @@
+%
+% cl_fonp02.m
+%
+% Control Law, Full Order, No Prediction
+%
+
+clear;
+
+T = 1;  % time step
+ncyl = 8;  % number of cylinders
+% stabilizes at rpm_lo then steps to rpm_hi
+rpm_hi = 700;  % idle rpm
+rpm_lo = 200;
+
+% Engine Model
+Gy = engine_model(T);
+[Phi, Gamma, H, J] = ssdata(Gy);
+n = length(Phi);  % order
+% engine model that outputs x instead of y
+Gx = ss(Phi, Gamma, eye(n), zeros(n,1), T);
+
+% Rz converts rpm to control (0-1)
+Rz = ss(0, 0, 0, [1/1450.1], T);
+
+% Hz gain, y = H*x
+Hz = ss(0, zeros(1,n), 0, H, T);
+
+% Find K
+% roots (arbitrary)
+%z1 = [(0.9 + 0.05i) (0.9 - 0.5i) 0];
+z1 = [(0.9 + 0.05i) (0.9 - 0.05i) 0.01];
+%z1 = [0.5 0.4 0];
+%z1 = [0.05 0.04 0];
+K1 = place(Phi, Gamma, z1);
+% create a gain of -K using D and zeroing others
+K1z = ss(zeros(1,1), zeros(1,n), zeros(1,1), -K1, T);
+
+% Build System
+% uncompensated feedback loop
+sys1 = feedback(Gx, K1z, [1], [1 2 3], +1);
+% add rpm to control gain to begining
+sys2 = series(Rz, sys1);
+% add y = H*x
+sys3 = series(sys2, Hz);
+sys3 = sminreal(sys3);
+
+% Input signal, rpm_lo then steps to rpm_hi
+Tend = 140;  % end in ticks
+c1 = Tend/T;
+sc = c1/2;  % point at which step occurs
+u1 = rpm_lo*ones([sc 1]);
+u2 = rpm_hi*ones([sc 1]);
+u = [u1; u2];
+
+% Simulate
+% initial x0
+x0 = rpm_lo*H;
+[y1,t1] = lsim1(sys3, u, [], x0);
+
+% be sure to avoid zeros, rpmtime doesn't handle them
+yX = y1;
+%yX = y2(10:end);  % all data, except first zeros (approx)
+%yX = y2((stepc+1):end);  % data after step
+rt = rpmtime(yX, ncyl);
[…]

### 2013-12-19 octave/cl_fonp03.m

@@ -1,0 +1,85 @@
+%
+% cl_fonp03.m
+%
+% Control Law, Full Order, No Prediction
+%
+
+clear;
+
+T = 1;  % time step
+ncyl = 8;  % number of cylinders
+% stabilizes at rpm_lo then steps to rpm_hi
+rpm_hi = 700;  % idle rpm
+rpm_lo = 200;
+
+% Engine Model
+Gy = engine_model(T);
+[Phi, Gamma, H, J] = ssdata(Gy);
+n = length(Phi);  % order
+% engine model that outputs x instead of y
+Gx = ss(Phi, Gamma, eye(n), zeros(n,1), T);
+
+% Rz converts rpm to control (0-1)
+Rz = ss(0, 0, 0, [1/1450.1], T);
+
+% Hz gain, y = H*x
+Hz = ss(0, zeros(1,n), 0, H, T);
+
+% Find K
+% roots (arbitrary)
+%z1 = [(0.9 + 0.05i) (0.9 - 0.5i) 0];
+z1 = [(0.9 + 0.05i) (0.9 - 0.05i) 0.01];
+%z1 = [0.5 0.4 0];
+%z1 = [0.05 0.04 0];
+K1 = place(Phi, Gamma, z1);
+% create a gain of -K using D and zeroing others
+K1z = ss(zeros(1,1), zeros(1,n), zeros(1,1), -K1, T);
+
+% Build System
+% uncompensated feedback loop
+sum1 = sumblk('x2 = x1 - rpm1');
+sum2 = sumblk('x2b = x1b - rpm1b');
+sum3 = sumblk('x2c = x1c - rpm1c');
+set(Gx, 'inname', 'e1', 'outname', {'x1', 'x1b', 'x1c'});
+sys1 = connect(sum1, sum2, sum3, Gx, {'e1', 'rpm1', 'rpm1b', 'rpm1c'}, {'x2', 'x2b', 'x2c'});
+sys2 = feedback(sys1, K1z, [1], [1 2 3], +1);
+set(sys2, 'inname', {'u2', 'rpm1', 'rpm1b', 'rpm1c'}, 'outname', {'x2', 'x2b', 'x2c'});
+set(Rz, 'inname', {'u1'}, 'outname', {'u2'});
+sys3 = connect(sys2, Rz, {'u1', 'rpm1', 'rpm1b', 'rpm1c'}, {'x2', 'x2b', 'x2c'});
+% add y = H*x
+sys4 = series(sys3, Hz);
[…]

### 2013-12-20 matlab/cd_plot2.m
applied Matlab compatibilies with Octave
Various changes which make the M-Files compatible
with both Matlab and Octave have been applied.
Octave is surprising compatible.  Only once case,
converting a z-domain transfor to state space,
hasn't been resolved.

- removed 'endfunction'
- removed 'endfor', replaced with 'end'
- Matlab does not support default function arguments,
  used 'narg' instead.
- Matlab accepts 'InputName' instead of 'inname' and
  'OutputName' instead of 'outname' for the control toolbox.

@@ -6,7 +6,7 @@ clear;
 
 T = 1;  % time step
 
-sys = tf([1 -1], [1 0], T, 'inname', 'y1', 'outname', 'u1');
+sys = tf([1 -1], [1 0], T, 'InputName', 'y1', 'OutputName', 'u1');
 
 u = [1 1 1 -3 -3 2 5 0];
 t = 0:(size(u,2)-1);  % start at zero

### 2013-12-20 matlab/cl_ff01.m

@@ -23,12 +23,12 @@ Gz = engine_model(T);
 % upper branch
 sum1 = sumblk('e1 = u + r1');
 % convert rpm to control
-Rz = tf([1], [1450.1], T, 'inname', 'e1', 'outname', 'x1');
-set(Gz, 'inname', 'x1', 'outname', 'x2');
+Rz = tf([1], [1450.1], T, 'InputName', 'e1', 'OutputName', 'x1');
+set(Gz, 'InputName', 'x1', 'OutputName', 'x2');
 sum2 = sumblk('y = x2 - r2');
 sys1 = connect(sum1, Rz, Gz, sum2, {'u', 'r1', 'r2'}, {'y'});
 % feedback branch
-Kz = tf([-K], [1], T, 'inname', 'y', 'outname', 'u');
+Kz = tf([-K], [1], T, 'InputName', 'y', 'OutputName', 'u');
 % full system
 sysX = feedback(sys1, Kz, [1], [1], +1);
 sysX = sminreal(sysX);

### 2013-12-20 matlab/cl_fonp03.m

@@ -40,11 +40,11 @@ K1z = ss(zeros(1,1), zeros(1,n), zeros(1,1), -K1, T);
 sum1 = sumblk('x2 = x1 - rpm1');
 sum2 = sumblk('x2b = x1b - rpm1b');
 sum3 = sumblk('x2c = x1c - rpm1c');
-set(Gx, 'inname', 'e1', 'outname', {'x1', 'x1b', 'x1c'});
+set(Gx, 'InputName', 'e1', 'OutputName', {'x1', 'x1b', 'x1c'});
 sys1 = connect(sum1, sum2, sum3, Gx, {'e1', 'rpm1', 'rpm1b', 'rpm1c'}, {'x2', 'x2b', 'x2c'});
 sys2 = feedback(sys1, K1z, [1], [1 2 3], +1);
-set(sys2, 'inname', {'u2', 'rpm1', 'rpm1b', 'rpm1c'}, 'outname', {'x2', 'x2b', 'x2c'});
-set(Rz, 'inname', {'u1'}, 'outname', {'u2'});
+set(sys2, 'InputName', {'u2', 'rpm1', 'rpm1b', 'rpm1c'}, 'OutputName', {'x2', 'x2b', 'x2c'});
+set(Rz, 'InputName', {'u1'}, 'OutputName', {'u2'});
 sys3 = connect(sys2, Rz, {'u1', 'rpm1', 'rpm1b', 'rpm1c'}, {'x2', 'x2b', 'x2c'});
 % add y = H*x
 sys4 = series(sys3, Hz);
@@ -82,4 +82,4 @@ title('Control Law, Full Feedback, No Prediction');
 xlabel('time (sec)');
 ylabel('rpm');
 
-print('cl_fonp02.eps', '-depsc2');
+print('cl_fonp03.eps', '-depsc2');

### 2013-12-20 matlab/dc_plot1.m

@@ -1,12 +1,16 @@
 %
-% dc_plot2.m
+% dc_plot.m
+%
+% Digital to steady state transfer function.
 %
 
 clear;
 
 T = 1;  % time step
 
-sys = tf([1 0], [1 -1], T, 'inname', 'u1', 'outname', 'y1');
+D1 = tf([1], [1 0], T, 'InputName', 'y1', 'OutputName', 'r1');
+S1 = sumblk('y1 = u1 + r1');
+sys = connect(D1, S1, 'u1', 'y1');
 
 u = [1 0 0 -4 0 5 3 3];
 t = 0:(size(u,2)-1);  % start at zero
@@ -15,8 +19,7 @@ figure;
 subplot(2,1,1);
 stairs(t,u);
 grid on;
-%axis auto;
-axis([0 7 -6 6]);
+axis auto;
 title('Input Signal');
 ylabel('u');
 
@@ -24,6 +27,5 @@ subplot(2,1,2);
 lsim(sys, u);
 grid on;
 axis auto;
-axis([0 7 -4 6]);
 
-print('dc_plot2.eps', '-depsc2');
+print('dc_plot1.eps', '-depsc2');

### 2013-12-20 matlab/dc_plot2.m

@@ -1,16 +1,12 @@
 %
-% dc_plot.m
-%
-% Digital to steady state transfer function.
+% dc_plot2.m
 %
 
 clear;
 
 T = 1;  % time step
 
-D1 = tf([1], [1 0], T, 'inname', 'y1', 'outname', 'r1');
-S1 = sumblk('y1 = u1 + r1');
-sys = connect(D1, S1, 'u1', 'y1');
+sys = tf([1 0], [1 -1], T, 'InputName', 'u1', 'OutputName', 'y1');
 
 u = [1 0 0 -4 0 5 3 3];
 t = 0:(size(u,2)-1);  % start at zero
@@ -19,7 +15,8 @@ figure;
 subplot(2,1,1);
 stairs(t,u);
 grid on;
-axis auto;
+%axis auto;
+axis([0 7 -6 6]);
 title('Input Signal');
 ylabel('u');
 
@@ -27,5 +24,6 @@ subplot(2,1,2);
 lsim(sys, u);
 grid on;
 axis auto;
+axis([0 7 -4 6]);
 
-print('dc_plot1.eps', '-depsc2');
+print('dc_plot2.eps', '-depsc2');

### 2013-12-20 matlab/lsim1.m

@@ -1,8 +1,16 @@
-function [y, t, x, i] = lsim1(Hz, u, t=[], x0=false);
+function [y, t, x, i] = lsim1(Hz, u, t, x0);
 % LSIM1 - Just like lsim excpt small values are removed.
 %
 % "small" means less than or equal to 10.
 %
+
+switch nargin
+    case 2
+        t = [];
+        x0=false;
+    case 3
+        x0=false;
+end
 
 [y, t, x] = lsim(Hz, u, t, x0);
 
@@ -12,7 +20,7 @@ for i = 1:length(y)
 	if (x > 10)
 		break;
 	end
-endfor
+end
 
 % adjust for removed values
 y = y(i:end);

### 2013-12-20 matlab/rpmtime.m

@@ -20,9 +20,7 @@ for i = 1:length(rpm)
 		ts(i) = inf;
 	else
 		ts(i) = 1/x;
-	endif
-endfor
+	end
+end
 
 ts = cumsum(ts);
-
-endfunction

### 2013-12-20 paper/main.tex
applied Matlab compatibilies with Octave
Various changes which make the M-Files compatible
with both Matlab and Octave have been applied.
Octave is surprising compatible.  Only once case,
converting a z-domain transfor to state space,
hasn't been resolved.

- removed 'endfunction'
- removed 'endfor', replaced with 'end'
- Matlab does not support default function arguments,
  used 'narg' instead.
- Matlab accepts 'InputName' instead of 'inname' and
  'OutputName' instead of 'outname' for the control toolbox.

@@
 \begin{center}
-\includegraphics[scale=0.7]{../octave/em_plot1}
+\includegraphics[scale=0.7]{../matlab/em_plot1}
 \end{center}
@@
 \begin{center}
-\includegraphics[scale=0.7]{../octave/em_plot2}
+\includegraphics[scale=0.7]{../matlab/em_plot2}
 \end{center}
@@
 \begin{center}
-\includegraphics[scale=0.6]{../octave/em_plot4}
+\includegraphics[scale=0.6]{../matlab/em_plot4}
 \end{center}
@@
 \begin{center}
-\includegraphics[scale=0.6]{../octave/dd_plot2}
+\includegraphics[scale=0.6]{../matlab/dd_plot2}
 \end{center}
@@
 \begin{center}
-\includegraphics[scale=0.6]{../octave/cl_ff01}
+\includegraphics[scale=0.6]{../matlab/cl_ff01}
 \end{center}
@@
 \begin{figure}[htbp!]
-\includegraphics[scale=0.7]{../octave/cd_plot1}
+\includegraphics[scale=0.7]{../matlab/cd_plot1}
 \caption{Response of full steady state input to delta output system
@@
 \begin{figure}[htbp!]
-\includegraphics[scale=0.7]{../octave/cd_plot2}
+\includegraphics[scale=0.7]{../matlab/cd_plot2}
 \caption{Response of simplified steady state input to delta output system
@@
 	label=lst:cd_plot1
-	]{../octave/cd_plot1.m}
+	]{../matlab/cd_plot1.m}
 
@@
 	label=lst:cd_plot2
-	]{../octave/cd_plot2.m}
+	]{../matlab/cd_plot2.m}
 
@@
 \begin{figure}[htbp!]
-\includegraphics[scale=0.7]{../octave/dc_plot2}
+\includegraphics[scale=0.7]{../matlab/dc_plot2}
 \caption{Response of delta input to steady state output system
@@
 	label=lst:dc_plot2
-	]{../octave/dc_plot2.m}
+	]{../matlab/dc_plot2.m}
 
@@
 \begin{center}
-\includegraphics[scale=0.5]{../octave/em_plot3}
+\includegraphics[scale=0.5]{../matlab/em_plot3}
 \end{center}
@@
 	label={lst:stable}
-	]{../octave/em_plot3.m}
+	]{../matlab/em_plot3.m}
 % }}}
@@
 	use this model.}
-	]{../octave/engine_model.m}
+	]{../matlab/engine_model.m}
 
@@
 		with no control. The $x$ axis is in ticks.}
-	]{../octave/em_plot1.m}
+	]{../matlab/em_plot1.m}
 
@@
 	with no control.  The $x$ axis is in time.}
-	]{../octave/em_plot2.m}
+	]{../matlab/em_plot2.m}
 
@@
 		versus time for a typical 600 to 1000 rpm range.}
-	]{../octave/em_plot4.m}
+	]{../matlab/em_plot4.m}
 % }}}
@@
 	caption={Matlab source of function to calculate a Sylvester Matrix.} 
-	]{../octave/sylvester.m}
+	]{../matlab/sylvester.m}
 
@@
 	caption={Matlab source of function to calculate zero time.} 
-	]{../octave/zerotime.m}
+	]{../matlab/zerotime.m}
 
@@
 	caption={Matlab source of modified lsim() function.} 
-	]{../octave/lsim1.m}
+	]{../matlab/lsim1.m}
 
@@
 		Direct Design.}
-	]{../octave/direct_design.m}
+	]{../matlab/direct_design.m}
 
@@
 	Direct Design compared to no control.}
-	]{../octave/dd_plot2.m}
+	]{../matlab/dd_plot2.m}
 
@@
 	caption={Matlab source used to plot the Control Law regulator.}
-	]{../octave/cl_ff01.m}
+	]{../matlab/cl_ff01.m}
 
@@
 Octave\autocite{octave}, an open source Matlab clone.
-When possible the examples were tested in Matlab/Simulink.
-However much of the code is dependent upon the
-Octave Control (CACSD) Tools\autocite{octave-control} which are
-not entirely compatible with Matlab.
+Except for a few instances (\verb+cl_fonp02.m+, \verb+cl_fonp03.m+) the
+M-Files are completely compatible between both Octave and Matlab.
+The entire control system is built within the M-File and has
+no dependence on Simulink.
+Octave uses the Octave Control (CACSD) Tools\autocite{octave-control} 
+which have equivalents in Matlab.
+In a few cases an equivalent Simulink diagram was built and in
+each of the cases the results agreed.
 This document and all of its code are released under the

### 2013-12-20 matlab/cd_plot1.m

@@ -6,8 +6,8 @@ clear;
 
 T = 1;  % time step
 
-D1 = tf([1], [1 0], T, 'inname', 'y1', 'outname', 'v1');
-D2 = tf([1], [1 0], T, 'inname', 'r1', 'outname', 'q1');
+D1 = tf([1], [1 0], T, 'InputName', 'y1', 'OutputName', 'v1');
+D2 = tf([1], [1 0], T, 'InputName', 'r1', 'OutputName', 'q1');
 sum1 = sumblk('y1 = u1 - r1');
 sum2 = sumblk('r1 = v1 + q1');
 sys = connect(D1, D2, sum1, sum2, 'u1', 'y1');

### 2013-12-20 matlab/cl_fonp02.m

@@ -31,7 +31,8 @@ Hz = ss(0, zeros(1,n), 0, H, T);
 z1 = [(0.9 + 0.05i) (0.9 - 0.05i) 0.01];
 %z1 = [0.5 0.4 0];
 %z1 = [0.05 0.04 0];
-K1 = place(Phi, Gamma, z1);
+%K1 = place(Phi, Gamma, z1);
+K1 = myacker(Phi, Gamma, z1);
 % create a gain of -K using D and zeroing others
 K1z = ss(zeros(1,1), zeros(1,n), zeros(1,1), -K1, T);
 

### 2013-12-20 matlab/em_plot3.m

@@ -20,7 +20,7 @@ for i = 1:length(us)
 
 	% save the stable value
 	y(i) = yp(end);	
-endfor
+end
 
 figure(1);
 

### 2013-12-20 matlab/cl_fonp03.m
applied Matlab specific changes
The .mdl files have been included.
And the workaround for the state space engine model
has been added.

@@ -13,8 +13,8 @@ rpm_hi = 700;  % idle rpm
 rpm_lo = 200;
 
 % Engine Model
-Gy = engine_model(T);
-[Phi, Gamma, H, J] = ssdata(Gy);
+[Phi, Gamma, H, J] = ss_engine_model(T);
+
 n = length(Phi);  % order
 % engine model that outputs x instead of y
 Gx = ss(Phi, Gamma, eye(n), zeros(n,1), T);

### 2013-12-20 matlab/ss_engine_model.m

@@ -1,0 +1,17 @@
+function [Phi, Gamma, H, J] = ss_engine_model(T);
+%  SS_ENGINE_MODEL
+%
+% State Space linear engine model with zero torque.
+%
+
+%T = 1;  % time step
+
+% For some reason Matlab doesn't convert from
+% the z-domain to the s-domain properly.
+% So here we just use what Octave generated.
+Phi = [0 0 -0.00237;
+	  -1 0 0.08442
+	   0 -10 1.83790];
+Gamma = [-4.33815; 0; 0];
+H = [0 0 -1];
+J = 0;

### 2013-12-20 paper/main.tex
added Colophon
A Colophon has been added that describes how this
document was designed.  The References section has
also moved to the end of the document.

@@
 inputs are constant.
-This paper shows how control system methods can be applied to
-the problem of idle stabilization.
+This paper investigates how control system designs can be applied
+to the problem of idle stabilization.
+Methods include: pole placement, direct design, and Control Law designs.
 Because the engine model is inherently discrete all of the methods
-used are also discrete.
-Methods include: pole placement, direct design, and
-various state space designs.
+discussed are also discrete.
 \end{abstract}
@@
 \section{Control Law, Intuitive Design}
+\label{sec:clid}
 
@@
 
-There are several requirements that must be met in order for this
-system to be stable.
+Several requirements must be satisfied in order for this system to be stable.
 First, the input to $G(z)$ must reflect the output without any offset.
-And second the system must stabilize at zero.
+Second, the system must stabilize at zero.
 These requirements can be verified intuitively.
@@
 The system diverges.
-This exercise can be repeated with a controller that has no offset at zero.
-The result is a negative feedback system that will stabilize at zero.
+This exercise can be repeated with a controller at zero with
+no input/output offset.
+The result is a stable negative feedback system.
 
@@
 satisfies neither of these requirements.
-The input does not reflect the output; the input range is from zero to
-one whereas the output range is from 200 to 1200.
-And operation at zero rpm is not realistic.
+There is an offset because the input has a range from zero to one which
+corresponds to a range of 200 to 1200 on the output.
+And operation at zero rpm is impossible.
 
@@
 
-The system in Figure \ref{fig:clid02} adds compensations to the
-engine model so that a regulator system can be constructed.
-An input of zero should produce the desired rpm.
-And this rpm input is converted to a control value ($R$).
+To resolve these issues compensation can be added to the
+system as shown in Figure \ref{fig:clid02}.
+An input of zero will produce the desired rpm.
+The rpm is then converted to a control value ($R$).
 And the reached rpm (output of $G(z)$) is subtracted from
@@
 
-Figure \ref{fig:cl_plot1} shows the response of this system.
-While the response isn't very good the design was trivial.
-A value for $K$ was arbitrarily chosen.
-Any values less than one were stable and larger values diverged.
-Later designs will improve performance by elaborating upon this
-system and developing more complex values for $K$.
-
-% }}}
-
+Figure \ref{fig:cl_plot1} shows the response of this system
+with an arbitrary value of $K$.
+Values less than or equal to one produced a stable response and
+anything larger will cause it to diverge.
+The response is mediocre but the design was trivial and
+required no calculations.
+Later designs will improve this system by constructing more
+complex values for $K$.
+
+% }}}
+
+% {{{ Control Law, Full Feedback, No Prediction
+\clearpage
+\section{Control Law, Full Feedback, No Prediction}
+\label{sec:clffnp}
+
+The simplest Control Law\autocite[Pg. 280]{franklin1998digital} design
+uses full feedback with no prediction.
+Ackermann's Formula is used to find $K$ for given values
+of $\Phi$ and $\Gamma$.
+Figure \ref{fig:clffnp01} shows the system.
+
+% {{{ fig:clffnp01
+\begin{figure}[!htp]
+\begin{center}
+
+\begin{tikzpicture}[>=triangle 60,
+	node distance=14mm, auto]
+
+	\node[sumxpxp] 	(sum1)	[]					{};
+	\node[gain]  	(R)		[right=of sum1] 	{$R$};
+	\node[gain]  	(Gz)	[right=of R] 		{$x(k + 1) = \Phi x + \Gamma u$};
+	\node[sumxpxn]	(sum2)	[right=of Gz]		{};
+	\node[gain]		(H)		[right=of sum2]		{$H$};
+	\node[gain] 	(K)		[below=of Gz] 		{$-K$};
+
+	\draw [<-] (sum1.north) -- ++(0,10mm) node[above]{$\text{rpm}$};
+	\draw [->] (sum1.east) -- (R.west);
+	\draw [->] (R.east) -- (Gz.west);
+	\draw [->] (Gz.east) -- (sum2.west);
+	\draw [<-] (sum2.north) -- ++(0,10mm) node[above]{$\text{rpm}$};
+	\draw [->] (sum2.east) -- ++(7mm,0) node[above,xshift=0mm]{$x$}
+		-| ++(0,-10mm)
+		|- (K.east);
+	\draw [->] (sum2.east) -- (H.west);
+	\draw [->] (H.east) -- ++(10mm,0) node[above,xshift=-3mm]{$y$};
+	\draw [<-] 
+		(sum1.west) -- ++(-10mm,0) node[above,xshift=3mm]{$u$}
+		|- (K.west);
+\end{tikzpicture}
+
+\end{center}
+
+\caption{Control Law system for full feedback with no prediction.
+The system is in the discrete state space form,
+with $\Phi$ and $\Gamma$, to allow Ackermann's formula to be
+used to find $K$.}
+\label{fig:clffnp01}
+\end{figure}
+% }}}
+
+Unfortunately the compensations that were added to make it possible
+to build a Control Law system (Section \ref{sec:clid}) create
+complications when applying Ackermann's Formula to find $K$.
+A system commonly given in the literature is
+shown in Figure \ref{fig:clffnp02}
+\footnote{Some diagrams omit $H$ but pass $x$ through an identity matrix.}.
+Here the scope of $\Phi$ and $\Gamma$ is clear and it is
+straight forward to apply Ackermann's Formula to find $K$.
+If the compensations in Figure \ref{fig:clffnp01} could be reduced
+to an equivalent transfer function with rpm set to a constant value
+Ackermann's could be equivalently applied.
+However the sum blocks cannot be reduced unless the constant input
+is zero, which is not the case here\footnote{Zero rpm is impossible, 700 rpm
+is a typical idle speed.}.
+
+% {{{ fig:clffnp02
+\begin{figure}[hpb!]
+\begin{center}
+
+\begin{tikzpicture}[>=triangle 60,
+	node distance=14mm, auto]
+
+	\node[gain]  	(Gz)	[] 		{$x(k + 1) = \Phi x + \Gamma u$};
+	\node[gain] 	(K)		[below=of Gz] 	{$-K$};
+	\node[gain]		(H)		[right=of Gz]	{$H$};
+
+	\draw [->]
+	 		(Gz.east) -- ++(7mm,0)
+			-| ++(0,-5mm)
+			|- (K.east);
+	\draw [->] (Gz.east) -- (H.west) node[above,xshift=-7mm]{$x$};
+	\draw [->] (H.east) -- ++(10mm,0) node[above,xshift=-3mm]{$y$};
+
+	\draw [<-]
+	 		(Gz.west) -- ++(-10mm,0)
+			-| ++(0,-10mm)
+			|- (K.west);
+\end{tikzpicture}
+
+\end{center}
+
+\caption{Typical Full Feedback system.}
+\label{fig:clffnp02}
+\end{figure}
+% }}}
+
+One possible solution is to rearrange the system as shown
+in Figure \ref{fig:clffnp03}.
+This resolves one of the sum blocks but another is still present.
+If it is assumed that this sum block can be ignored the following
+values were found\footnote{Matlab source is given in
+Listing \ref{lst:cl_fonp03} of Appendix \ref{app:cl_matlab}}.
+The resulting response is shown in Figure \ref{fig:cl_fonp03}.
+
+\begin{align*}
+	\text{roots} :& \quad (0.09 \pm 0.05i) \\
+				& \quad 0.01 \\
+	K &= [ -6.4313\mbox{\sc{e}-}03 \quad 8.6624\mbox{\sc{e}-}03
+			\quad -3.1446\mbox{\sc{e}-}04 ]
+\end{align*}
+
+% {{{ fig:clffnp03
+\begin{figure}[hpb!]
+\begin{center}
+
+\begin{tikzpicture}[>=triangle 60,
+	node distance=14mm, auto]
+
+	\node[gain]  	(R)		[] 				{$R$};
+	\node[sumpp]	(sum1)	[right=of R]	{};
+	\node[gain]  	(Gz)	[right=of sum1]	{$x(k + 1) = \Phi x + \Gamma u$};
+	\node[sumxpxn]	(sum2)	[right=of Gz]	{};
+	\node[gain] 	(K)		[below=of Gz] 	{$-K$};
+	\node[gain]		(H)		[right=of sum2]	{$H$};
+
+	\draw [->] ++(-15mm,0) node[left]{$\text{rpm}$} -- (R.west);
+	\draw [->] (R.east) -- (sum1.west);
+	\draw [->] (sum1.east) -- (Gz.west);
+	\draw [<-] (sum2.north) -- ++(0,10mm) node[above]{$\text{rpm}$};
+	\draw [->] (Gz.east) -- (sum2.west);
+	\draw [->]
+	 		(sum2.east) -- ++(7mm,0)
+			-| ++(0,-5mm)
+			|- (K.east);
+	\draw [->] (sum2.east) -- (H.west) node[above,xshift=-7mm]{$x$};
+	\draw [->] (H.east) -- ++(10mm,0) node[above,xshift=-3mm]{$y$};
+	\draw [<-]
+	 		(sum1.south)
+			|- (K.west);
+\end{tikzpicture}
+
+\end{center}
+
+\caption{Rearranged Control Law system to exclude one sum block.}
+\label{fig:clffnp03}
+\end{figure}
+% }}}
+
+% {{{ fig:cl_fonp03
+\begin{figure}[htbp!]
+\begin{center}
+\includegraphics[scale=0.7]{../matlab/cl_fonp03}
+\end{center}
+\caption{Output response if sum block is ignored and $K$ is found.
+Rpm input is 200 until 0.6 sec where it steps to 700.
+The output should reflect the input which this clearly does not.
+Matlab source is given in Listing \ref{lst:cl_fonp03}
+of Appendix \ref{app:cl_matlab}.
+}
+\label{fig:cl_fonp03}
+\end{figure}
+% }}}
+
+It is clear from the response that the sum block plays an important
+role with regard to the calculation of $K$ using Ackermann's Formula.
+However it has not been determined how this problem can be
+resolved to find a valid value.
+
+% }}}
+
+% {{{ Conclusion
 \clearpage
@@
 
-% TODO
+The goal of this paper was to apply a range of control system
+designs to the problem of idle stabilization with a spark ignition
+engine.
+Unfortunately, numerous difficulties were encountered with this
+real world problem.
+
+The engine model took a torque input to a sum block.
+To simplify calculations it was assumed to be zero which allowed
+the simplification of the system in to a single transfer function.
+A non-zero value would have substantially complicated matters.
+
+The engine model did not accept steady state inputs.
+Transfer functions had to be found for converting a steady state to a
+delta and from a delta to a steady state.
+
+The engine model cannot operate at zero rpm because this is a
+nonsensical value.
+Similarly a unit step input could not be used because both zero
+and one rpm are equally nonsensical.
+The scheme that was devised was to apply a custom input that started
+at 200 rpm and then stepped to 700 rpm.
+Then the resulting response could be seen under a typical range.
+
+Several designs were successfully applied.
+A direct design using pole placement was constructed that had improved
+performance compared to no control
+And a Control Law design built using an intuitive design was a success.
+A Control Law with Full Feedback was attempted but was not a success.
+The complications created by a sum block prevented a correct $K$
+from being found using Ackermann's Formula.
+No designs beyond this were attempted because they would be equally
+hampered by the same problems.
+
+% }}}
 
@@
 
+\lstinputlisting[
+	caption={Matlab source of rpmtime() function.} 
+	]{../matlab/rpmtime.m}
+
 % }}}
@@
 
-Matlab source code used to the Control Law regulator.
+The following Matlab source code was using for the Control Law Designs
+(Section \ref{sec:clid}, \ref{sec:clffnp}).
 
@@
 	]{../matlab/cl_ff01.m}
+
+\lstinputlisting[
+	caption={Matlab source used to plot the Control Law with no predicition
+	if sum block is ignored (Figure \ref{fig:cl_fonp03}).},
+	label=lst:cl_fonp03,
+	]{../matlab/cl_fonp03.m}
 
@@
 GNU General Public License and freely
-available\autocite{github_jmahler_idle-control}.
+available to download and modify\autocite{github_jmahler_idle-control}.
 \nocite{github_jmahler_control}
